---
title: "Final_phase"
format: pdf
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

# Library import

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr)
library(maps)
library(viridis)
library(ggrepel)
library(broom)
library(scales)
library(patchwork)
library(ggthemes)
library(lmtest) #heteroscedasticity 

# Data import

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
                           delim = ";", escape_double = FALSE, trim_ws = TRUE)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv", skip = 3)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv", skip = 3)

Capital_formation <- read_csv("Database_project/Capital_formation.csv", skip = 3
)

Population <- read_csv("Database_project/Population.csv", skip = 3)


# Cleaning billionaires data set

useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |> # standardization of names
  rename(COUNTRY = citizenship)

Billionaires_1996 <-
  Billionaires |>
  filter(year == "1996")

Billionaires_2001 <-
  Billionaires |>
  filter(year == "2001")

Billionaires_2014 <-
  Billionaires |>
  filter(year == "2014")

Billionaires_wide <-
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Cleaning the IMF dataset

Country <-
  Billionaires |>
  distinct(COUNTRY)

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- c(
  "Bahrain, Kingdom of" = "Bahrain",
  "China, People's Republic of" = "China",
  "Egypt, Arab Republic of" = "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" = "Hong Kong",
  "Kazakhstan, Republic of" = "Kazakhstan",
  "Liechtenstein, Principality of" = "Liechtenstein",
  "Lithuania, Republic of" = "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" = "Macau",
  "Netherlands, The" = "Netherlands",
  "Poland, Republic of" = "Poland",
  "Russian Federation" = "Russia",
  "Korea, Republic of" = "South Korea",
  "Eswatini, Kingdom of" = "Swaziland",
  "Taiwan Province of China" = "Taiwan",
  "Tanzania, United Republic of" = "Tanzania",
  "Türkiye, Republic of" = "Turkey",
  "Venezuela, República Bolivariana de" = "Venezuela"
)



IMF <-
  IMF |>
  mutate(COUNTRY = recode(COUNTRY, !!!map_IMF))

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Cleaning the WB datasets

WB_names <-
  FDI |>
  distinct(`Country Name`)

map_WB <- c(
  "Czechia" = "Czech Republic",
  "Egypt, Arab Rep." = "Egypt",
  "Hong Kong SAR, China" = "Hong Kong",
  "Macao SAR, China" = "Macau",
  "Russian Federation" = "Russia",
  "Korea, Rep." = "South Korea",
  "Venezuela, RB" = "Venezuela",
  "Viet Nam" = "Vietnam"
)

FDI <-
  FDI |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Population <-
  Population |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)

```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) State of the world of billionaires in 2014

### 1.1 Global mapping of billionaires

```{r}
#| message: false
#| warning: false
#| echo: false

# Aggregation by country: number and total wealth

billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data

world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format

country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
    country_mapping[COUNTRY], # Apply mapping if found
    COUNTRY
  )) # Keep original name otherwise

# Join billionaire data with map data

world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country

map_count <- ggplot() +
  geom_polygon(
    data = world_map, # Draw all countries as base layer
    aes(x = long, y = lat, group = group), # Define coordinates and grouping
    fill = "grey90", color = "white", linewidth = 0.1
  ) + # Draw only countries with billionaires
  geom_polygon(
    data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
    aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
    color = "white", linewidth = 0.1
  ) +
  scale_fill_viridis(
    option = "plasma",
    name = "Number of\nbillionaires",
    trans = "log10", # Apply logarithmic transformation
    breaks = c(1, 5, 10, 50, 100, 500),
    direction = -1
  ) +
  theme_void() + # Remove axes and background for clean map
  labs(title = "Global distribution of billionaires in 2014") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```
For this specific map, we used AI assistance to help refine the R code and generate a clearer, more visual representation. We already had the initial intuition and the base approach, but we chose this mapping option because it is easier to read and more informative than a standard “classic” ggplot output.

This map shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs.


### 1.2 Caracteristics and origin of billionaires' wealth 

```{r}
#| message: false
#| warning: false
#| echo: false

Sectors <-
  Billionaires_2014 |>
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |>
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(
    cols = c(nb_self_made, nb_inherited),
    names_to = "origine",
    values_to = "count"
  ) |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Distribution of billionaires by sector and origin of wealth in 2014",
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  ) +
  theme_clean(base_size = 10)
```

This first graph illustrates the distribution of billionaires across various industries, highlighting that real estate is the most prominent sector for amassing such wealth, followed by retail and construction. A significant trend across almost all sectors is the clear dominance of self-made individuals over those who inherited their fortunes. This disparity is particularly pronounced in modern industries like technology, software, and hedge funds, where inherited wealth is virtually non-existent, whereas more traditional sectors like oil and banking show a relatively higher presence of generational wealth.


```{r}
#| message: false
#| warning: false
#| echo: false


Source <-
  Billionaires |>
  drop_na(selfmade) |>
  summarise(n = n(), .by = c(selfmade, year))

ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(
    title = "Source of wealth by year",
    y = "Year",
    x = "Number",
    fill = "Origin"
  ) +
  theme_clean(base_size = 10)
```

This graph tracks the evolution of billionaire wealth origins from 1996 to 2014, revealing a dramatic surge in the total number of billionaires globally, which increased from approximately 400 to over 1,600. While the billionaire population was much smaller and more evenly divided between inherited and self-made origins in 1996, the landscape shifted significantly by 2014. The expansion of this demographic over the eighteen-year period has been overwhelmingly driven by self-made entrepreneurs, who now constitute the vast majority of the total, indicating a global shift toward new wealth creation rather than the simple maintenance of inherited estates.

## 2) First model

```{r}
#| label: First model
#| message: false
#| warning: false
#| echo: false

# Data preparation

reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |>
              select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |>
              select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |>
              select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |>
              select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

# First model

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)
summary(modele)


```

Our baseline regression links the (log) number of billionaires in a country to our taxation proxy (tax revenue), economic size (GDP), openness (FDI), and investment capacity (capital formation). Overall, the model fits the data well (Adjusted R²= 0.771; F-test p-value < 10^(-14)), suggesting these macroeconomic factors explain a large share of cross-country variation in billionaire presence.

The coefficient on tax revenue is positive (β = 0.022) and marginally significant at the 10% level (p = 0.097). Interpreting tax revenue as a proxy for the overall level of taxation/fiscal pressure, this result suggests that countries with stronger tax capacity tend to host more, not fewer, billionaires. Economically, this is consistent with the idea that higher-tax countries often also provide the institutional and public-good environment that supports wealth creation and wealth retention—stable legal systems, infrastructure, education, and high-quality public services.

Therefore, taxation does not automatically translate into a weaker ability to attract or maintain top wealth holders. In the context of an open economy, this weak but positive association also suggests that “competitiveness” is not solely driven by low taxation: some high-tax jurisdictions can remain attractive because they compensate with stability and market opportunities.

The strongest determinant is economic scale: log GDP is highly significant and large. This implies that larger economies systematically host more billionaires, which aligns with the idea that big markets generate more high-value firms, deeper capital markets, and more opportunities for large fortunes. This is important for our research question because it highlights that billionaire wealth concentration is strongly linked to market size and overall economic performance—taxation effects must be interpreted conditional on these structural fundamentals.

FDI is also positive and significant, indicating that more open and internationally connected economies tend to have more billionaires. This supports the “national competitiveness in an open economy” channel: countries that attract foreign investment and integrate into global capital flows are also those where billionaire wealth is more prevalent. In other words, openness and investment attractiveness appear complementary to the presence of ultra-wealthy individuals rather than substitutable.

Finally, capital formation is not statistically significant (p = 0.40), suggesting that once GDP size and FDI are accounted for, differences in domestic investment intensity (as measured here) do not add much explanatory power for billionaire counts.

Taken together, this baseline model suggests that, using tax revenue as a proxy for taxation, higher taxation is not associated with a lower billionaire presence, while economic performance (GDP) and openness/competitiveness (FDI) are strongly associated with billionaire concentration. This supports an interpretation where taxation can coexist with competitiveness in open economies, provided the broader institutional and market conditions remain favorable.

### 2.1 Multicolinearity analysis

We here test for multicollinearity among the explanatory variables using the Variance Inflation Factor (VIF).

```{r}
#| label: Multicolinearity analysis
#| message: false
#| warning: false
#| echo: false
library(car) #vif, setting it here to avoid conflict with dplyr 
vif_df <- data.frame(
  Variable = names(vif(modele)),
  VIF = as.numeric(vif(modele))
)  
vif_df
```

The VIF values (≈1.19–1.34) are close to 1, indicating no meaningful multicollinearity among the regressors, so coefficient significance is unlikely to be driven by correlated explanatory variables.


### 2.2 Normality test for residuals

Here, we test whether our data are distributed according to a normal distribution. To do this, we calculate the Shapiro-Wilk normality test statistics. We obtain the following results :

```{r}
#| label: Normality test for residuals
#| message: false
#| warning: false
#| echo: false

shapiro_test <- shapiro.test(residuals(modele))

shapiro_test

```

- The p-value is 0.7615 :  It is well above the critical threshold of 0.05 (or 5%), which means that we do not reject the null hypothesis (H0). We conclude that the residuals of our model follow a normal distribution.
- The points (our data) follow the reference line almost perfectly, there are no major deviations, either at the extremes (no serious outliers) or in the centre. The distribution is symmetrical.

### 2.3 Heteroscedasticity 

We here test for heteroscedasticity using the Breusch–Pagan test to check whether the error variance is constant. 

```{r}
#| label: Heteroscedasticy 
#| message: false
#| warning: false
#| echo: false

bp<-bptest(modele)


bp_df <- data.frame(
  Test = "Studentized Breusch–Pagan",
  Test_statistic = as.numeric(bp$statistic),
  p_value = as.numeric(bp$p.value)
)

bp_df
```

With a p-value of 0.7439, we fail to reject the null hypothesis of constant error variance, meaning there is no evidence of heteroskedasticity in the model.

### 2.4 Conclusion
Overall, the diagnostic tests support the validity of the OLS assumptions for our first model.
The VIF results (around 1.2–1.3) indicate no multicollinearity, so the regressors are not strongly correlated.
The Shapiro–Wilk test (p = 0.7615) suggests the residuals are approximately normally distributed.
The Breusch–Pagan test (p = 0.7439) shows no evidence of heteroskedasticity, meaning the error variance can be treated as constant.
Taken together, these checks imply that our coefficient estimates and standard inference from this model are generally reliable.


## 3) MODÉLISATION DE LA FUITE FISCALE DES MILLIARDAIRES
```{r}

# 1) CONSTRUCTION DU SCORE DE RISQUE DE FUITE FISCALE
# OBJECTIF : Créer un score de 0 à 1 qui mesure la probabilité qu'un 
# milliardaire quitte son pays si une taxe est instaurée
# Basé sur 4 critères : taxation, secteur mobile, self-made, proximité refuge

## CHUNK 1.1 : Définition des paramètres de base
# -----------------------------------------------------------------------------
# On identifie quels secteurs sont mobiles (tech, finance) et quels pays 
# sont des refuges fiscaux reconnus

# Secteurs considérés comme "mobiles" (faciles à déplacer)
<<<<<<< HEAD
mobile_sectors <- c("Diversified financial", "Technology-Computer", "Hedge funds", "Private equity/leverage buyout",
                    "Venture capital", "Media", "Money Management")

# Liste des paradis/refuges fiscaux
tax_havens <- c("Switzerland", "Singapore", "Hong Kong", "Liechtenstein",
                "Monaco", "Luxembourg", "United Arab Emirates", "Bermuda", "Ireland", "Guernsey")
=======
mobile_sectors <- c("Finance", "Technology", "Hedge funds", "Private equity",
                    "Money Management", "Investments")

# Liste des paradis/refuges fiscaux
tax_havens <- c("Switzerland", "Singapore", "Hong Kong", "Liechtenstein",
                "Monaco", "Luxembourg", "United Arab Emirates", "Bermuda")
>>>>>>> 019498fb0ba5142bf538ecc7a70a4f96b4b3edbf

```

```{r}
## CHUNK 1.2 : Calcul des 4 composantes du score
# -----------------------------------------------------------------------------
# Pour chaque milliardaire, on calcule 4 indicateurs qui seront ensuite
# combinés pour obtenir le score final de risque de fuite

flight_data <- Billionaires_2014 %>%
  left_join(Tax_revenue_reduced %>% select(COUNTRY, tax = `2014`), by = "COUNTRY") %>%
  filter(!is.na(tax)) %>%
  mutate(
    # COMPOSANTE 1 : Pression fiscale (0 à 1)
    # Plus la taxation est élevée, plus le risque de fuite est grand
    # On normalise entre 0 (taxation minimale) et 1 (taxation maximale)
    tax_normalized = (tax - min(tax, na.rm = TRUE)) / 
                     (max(tax, na.rm = TRUE) - min(tax, na.rm = TRUE)),
    
    # COMPOSANTE 2 : Secteur mobile (0 ou 1)
    # = 1 si le milliardaire est dans tech/finance (facile à déplacer)
    # = 0 sinon (ex: industrie, immobilier = moins mobile)
    mobile_sector = ifelse(industry %in% mobile_sectors, 1, 0),
    
    # COMPOSANTE 3 : Self-made (0 ou 1)
    # = 1 si self-made (plus mobile, entrepreneurial)
    # = 0 si héritier (attachement familial au pays)
    is_selfmade = ifelse(selfmade == 1, 1, 0),
    
    # COMPOSANTE 4 : Proximité d'un refuge fiscal (0 ou 1)
    # = 1 si le pays est dans la même région qu'un refuge
    # = 0 sinon
    # On considère que les Européens ont accès à la Suisse,
    # les Asiatiques à Singapour/HK, etc.
<<<<<<< HEAD
    
    
    near_haven = case_when(
      # Europe : proche de Suisse, Monaco, Luxembourg
      COUNTRY %in% c("France", "Germany", "Italy", "Spain", "United Kingdom", 
                     "Belgium", "Netherlands", "Austria", "Sweden", "Norway", 
                     "Portugal", "Denmark", "Czech Republic", "Finland", "Greece",
                     "Poland", "Romania", "Cyprus", "Lithuania") ~ 1,
      # Asie : proche de Singapour, Hong Kong
      COUNTRY %in% c("China", "Japan", "South Korea", "India", "Malaysia",
                     "Thailand", "Indonesia", "Taiwan", "Philippines", "Macau", "Vietnam") ~ 1,
      # Moyen-Orient : proche des Émirats
      COUNTRY %in% c("Saudi Arabia", "Kuwait", "Israel", "Turkey", "Egypt", "Lebanon", "Oman") ~ 1,
      # Amérique du Nord : proche des Bermudes, Caïmans (via accès facile)
      COUNTRY %in% c("United States", "Canada", "Brazil", "Mexico", "Colombia", "Venezuela", "St. Kitts and Nevis") ~ 1,
=======
    near_haven = case_when(
      # Europe : proche de Suisse, Monaco, Luxembourg
      COUNTRY %in% c("France", "Germany", "Italy", "Spain", "United Kingdom", 
                     "Belgium", "Netherlands", "Austria", "Sweden", "Norway") ~ 1,
      # Asie : proche de Singapour, Hong Kong
      COUNTRY %in% c("China", "Japan", "South Korea", "India", "Malaysia",
                     "Thailand", "Indonesia", "Taiwan", "Philippines") ~ 1,
      # Moyen-Orient : proche des Émirats
      COUNTRY %in% c("Saudi Arabia", "Kuwait", "Israel", "Turkey", "Egypt") ~ 1,
      # Amérique du Nord : proche des Bermudes, Caïmans (via accès facile)
      COUNTRY %in% c("United States", "Canada") ~ 1,
>>>>>>> 019498fb0ba5142bf538ecc7a70a4f96b4b3edbf
      TRUE ~ 0
    ),
    
    # Si le milliardaire est DÉJÀ dans un refuge fiscal, son risque = 0
    # (il n'a pas besoin de fuir, il est déjà là où c'est avantageux)
    in_tax_haven = ifelse(COUNTRY %in% tax_havens, 1, 0)
  )

```

```{r}
## CHUNK 1.3 : Calcul du score final pondéré
# -----------------------------------------------------------------------------
# On combine les 4 composantes avec des poids pour obtenir un score final
# Poids choisis selon l'importance supposée de chaque facteur

flight_data <- flight_data %>%
  mutate(
    # Score = somme pondérée des 4 composantes
    # Taxation = 40% (facteur principal)
    # Secteur mobile = 25%
    # Self-made = 20%
    # Proximité refuge = 15%
<<<<<<< HEAD
    flight_risk_score = (tax_normalized * 0.45 + 
                        mobile_sector * 0.30 + 
                        is_selfmade * 0.15 + 
                        near_haven * 0.10),
=======
    flight_risk_score = (tax_normalized * 0.40 + 
                        mobile_sector * 0.25 + 
                        is_selfmade * 0.20 + 
                        near_haven * 0.15),
>>>>>>> 019498fb0ba5142bf538ecc7a70a4f96b4b3edbf
    
    # Ajustement : si déjà dans un refuge, score = 0
    flight_risk_score = ifelse(in_tax_haven == 1, 0, flight_risk_score),
    
    # Catégorisation pour faciliter l'analyse
    risk_category = case_when(
      flight_risk_score < 0.3 ~ "Faible",
      flight_risk_score >= 0.3 & flight_risk_score < 0.6 ~ "Moyen",
      flight_risk_score >= 0.6 ~ "Élevé"
    )
  )

```

```{r}
## CHUNK 1.4 : Agrégation par pays
# -----------------------------------------------------------------------------
# On calcule pour chaque pays le risque moyen de fuite et le nombre
# de milliardaires à haut risque

country_risk <- flight_data %>%
  group_by(COUNTRY) %>%
  summarise(
    nb_billionaires = n(),
    total_wealth = sum(networthusbillion, na.rm = TRUE),
    avg_flight_risk = mean(flight_risk_score, na.rm = TRUE),
    nb_high_risk = sum(risk_category == "Élevé", na.rm = TRUE),
    pct_high_risk = (nb_high_risk / n()) * 100,
    tax = mean(tax, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_flight_risk))

<<<<<<< HEAD
#ponderer par nb de milliardaires par pays à voir 

=======
>>>>>>> 019498fb0ba5142bf538ecc7a70a4f96b4b3edbf
# Affichage des pays les plus vulnérables
cat("=== TOP 10 PAYS LES PLUS VULNÉRABLES À LA FUITE FISCALE ===\n")
print(head(country_risk, 10))
```

