---
title: "Taxing the Titans : An Econometric Study of Billionaire Wealth Drivers"
subtitle: "Université Paris-Dauphine PSL"
author:
  - name: Philipine Poirier
  - name: Tony De Amorim
date: today
date-format: "MMMM D, YYYY"
abstract: |
  This project was carried out as part of the data management course taught during the first year of the Quantitative Economics master's programme and International Affairs and Development master's programme. AI is indicated each time it was used and was only employed to correct the code, confirm intuitions and expand on our initial ideas. 
format:
  pdf:
    geometry:
      - left=20mm   
      - right=20mm  
      - top=25mm    
      - bottom=25mm 
    papersize: a4
    toc: true
    number-sections: true
bibliography: Ressources/references.bib
csl: Ressources/apa.csl
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

# Library import

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr) # working with df
library(maps) # use of mpa objects at multiple time
library(viridis) # for better color scale
library(ggrepel)
library(broom)
library(scales)
library(patchwork) # for better graph display
library(ggthemes) # for better graph display
library(lmtest) # heteroscedasticity
library(knitr) # for better df display
library(kableExtra) # for better df display
library(modelsummary)

# Data import

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv", skip = 3)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv", skip = 3)

Capital_formation <- read_csv("Database_project/Capital_formation.csv", skip = 3)

Population <- read_csv("Database_project/Population.csv", skip = 3)


# Cleaning billionaires data set

useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |> # standardization of names
  rename(COUNTRY = citizenship)

Billionaires_1996 <-
  Billionaires |>
  filter(year == "1996")

Billionaires_2001 <-
  Billionaires |>
  filter(year == "2001")

Billionaires_2014 <-
  Billionaires |>
  filter(year == "2014")

Billionaires_wide <-
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Cleaning the IMF dataset

Country <-
  Billionaires |>
  distinct(COUNTRY)

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- list(
  "Bahrain, Kingdom of" ~ "Bahrain",
  "China, People's Republic of" ~ "China",
  "Egypt, Arab Republic of" ~ "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" ~ "Hong Kong",
  "Kazakhstan, Republic of" ~ "Kazakhstan",
  "Liechtenstein, Principality of" ~ "Liechtenstein",
  "Lithuania, Republic of" ~ "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" ~ "Macau",
  "Netherlands, The" ~ "Netherlands",
  "Poland, Republic of" ~ "Poland",
  "Russian Federation" ~ "Russia",
  "Korea, Republic of" ~ "South Korea",
  "Eswatini, Kingdom of" ~ "Swaziland",
  "Taiwan Province of China" ~ "Taiwan",
  "Tanzania, United Republic of" ~ "Tanzania",
  "Türkiye, Republic of" ~ "Turkey",
  "Venezuela, República Bolivariana de" ~ "Venezuela"
)



IMF <-
  IMF |>
  mutate(COUNTRY = case_match(COUNTRY, !!!map_IMF, .default = COUNTRY))

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Cleaning the WB datasets

WB_names <-
  FDI |>
  distinct(`Country Name`)

map_WB <- list(
  "Czech Republic" ~ "Czechia",
  "Egypt, Arab Rep." ~ "Egypt",
  "Hong Kong SAR, China" ~ "Hong Kong",
  "Macao SAR, China" ~ "Macau",
  "Russian Federation" ~ "Russia",
  "Korea, Rep." ~ "South Korea",
  "Venezuela, RB" ~ "Venezuela",
  "Viet Nam" ~ "Vietnam"
)

FDI <-
  FDI |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Population <-
  Population |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Setting the color theme for our graphs

theme_set(theme_clean(base_size = 8)) 
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill   = "viridis",
  ggplot2.discrete.colour   = scales::viridis_pal(option = "D")(8),
  ggplot2.discrete.fill     = scales::viridis_pal(option = "D")(8)
)

couleur_viridis <- scales::viridis_pal(option = "D")(8)[1] 

update_geom_defaults("col",   list(fill = couleur_viridis, colour = "white"))
update_geom_defaults("bar",   list(fill = couleur_viridis, colour = "white"))
update_geom_defaults("point", list(colour = couleur_viridis))
update_geom_defaults("line",  list(colour = couleur_viridis))

```

\newpage

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade in an era of unprecedented wealth concentration. Indeed, the top 0.001% owns today three times more wealth than the entire bottom half of humanity combined[^World_inequality].In June 2024, French economist Gabriel Zucman presented to the G20 a proposal[^Zucman_paper] for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services, opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension-with public debt exceeding 115% of GDP[^French_debt] and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.

Our study therefore aims to gain a better understanding of the key characteristics of the wealthiest individuals and to improve our ability to predict their tax behavior After documenting key patterns in billionaire characteristics, we construct an individual flight-risk index based on sector portability, the origin of wealth (self-made versus inherited), and proximity to low-tax jurisdictions. Rather than imposing an exogenous “flight rate,” we model it as an outcome of observed determinants and translate the index into relocation probabilities using a calibrated logistic specification. This logistic modeling revels that the risk of mobility is highly concentrated among self-made billionaires operating in dematerialized sectors such as tech or finance.

Finally, our revenue simulations for a 2% tax demonstrate a critical trade-off:

- Unilateral implementation by France would result in a 18.7% loss of revenue due to capital flight.

- International coordination (G20 or quasi-global) automatically reduces relocation options, bringing the leakage rate down to 4% and increasing collection efficiency to 95.9%.

The study concludes that international tax cooperation is a prerequisite for transforming tax justice into an effective budgetary lever without compromising national competitiveness.

## Database description

A more detailed presentation of the data used in this project can be find on the [github repository](https://github.com/TonyDeAmorim/Taxing-the-Titans-An-Econometric-Study-of-Billionaire-Wealth-Drivers) of the project.

- *The origins of the superrich: The billionaire characteristics database* [@Peterson2016] : A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- GDP in PPP from the IMF data explorer[^IMF]: This variable captures the overall economic performance and living standards of each country, adjusted for price level differences. Since wealth creation and the emergence of billionaires are closely linked to the size, productivity and income level of national economies, PPP-adjusted GDP is a natural control variable to account for cross-country differences in economic capacity.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and Tax revenue as a % of GDP, used to control for market scale, investment climate, and to approximate wealth tax.

All the World Bank's datasets used are listed here :

-   The [Total population](https://data.worldbank.org/indicator/SP.POP.TOTL) dataset
-   The [Gross capital formation (% of GDP)](https://data.worldbank.org/indicator/NE.GDI.TOTL.ZS) dataset
-   The [Foreign direct investment, net inflows (% of GDP)](https://data.worldbank.org/indicator/BX.KLT.DINV.WD.GD.ZS) dataset
-   The [Tax revenue (% of GDP)](https://data.worldbank.org/indicator/GC.TAX.TOTL.GD.ZS) dataset



# History and overview of the world's wealthiest individuals

In this section, we step back from the policy question to situate billionaire wealth in its broader context. The aim is to provide a brief descriptive overview of billionaire wealth over time and its cross-sectional profile in 2014, focusing on trends in aggregate net worth, the rise of self-made fortunes, and their geographic and sectoral concentration.

## The evolution of billionaires wealth over the years 

```{r}
#| label: fig-wealth-evolution
#| fig-cap: "Evolution of wealth through the years"
#| echo: false
#| message: false
#| warning: false

Total_wealth <-
  Billionaires |> 
  summarise(total_wealth = sum(networthusbillion), .by = year) |> 
  mutate(year = factor(year))

ggplot(Total_wealth, aes(x = year, y =total_wealth))+
  geom_col()+
  labs(x = "Year", y = "Total net worth in USD billions")

```

@fig-wealth-evolution shows the evolution of the aggregate wealth of the largest fortunes between 1996, 2001 and 2014. There has been a steady increase, but particularly rapid since the beginning of the 21st century: +261% between 2001 and 2014 compared to +70% between 1996 and 2001. This difference can be explained by the emergence of information and communication technologies at the beginning of the century, the creation of large-scale oligopolies through the proliferation of M&A transactions, and the ongoing process of globalization. These phenomena, combined with very high returns on capital during this period, even despite the crises, explain the results shown in this figure. 

```{r}
#| label: fig-wealth-source-evolution
#| fig-cap: "Evolution of the origin of wealth through the years"
#| echo: false
#| message: false
#| warning: false

Source <-
  Billionaires |>
  drop_na(selfmade) |>
  summarise(n = n(), .by = c(selfmade, year))

ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(
    title = "Source of wealth by year",
    y = "Year",
    x = "Number",
    fill = "Origin"
  )
```


As @fig-wealth-source-evolution indicates a surge in the number of billionaires occurred from 1996 to 2014, which increased from approximately 400 to over 1,600. While the billionaire population was much smaller and more evenly divided between inherited and self-made origins in 1996, the landscape shifted significantly by 2014. The expansion of this demographic over the eighteen-year period has been overwhelmingly driven by self-made entrepreneurs, who now constitute the vast majority of the total, indicating a global shift toward new wealth creation rather than the simple maintenance of inherited estates.


## State of the world of billionaires in 2014

@fig-Mapping-graph-of-billionaires shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs. This visualization thus confirmed our intuition to econometrically link billionaires and national macroeconomic characteristics.

```{r}
#| label: fig-Mapping-graph-of-billionaires
#| fig-cap: "Mapping of billionaires in the world in 2014"
#| echo: false
#| message: false
#| warning: false

# Aggregation by country: number and total wealth

billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data

world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format

country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
    country_mapping[COUNTRY], # Apply mapping if found
    COUNTRY
  )) # Keep original name otherwise

# Join billionaire data with map data

world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country

map_count <- ggplot() +
  geom_polygon(
    data = world_map, # Draw all countries as base layer
    aes(x = long, y = lat, group = group), # Define coordinates and grouping
    fill = "grey90", color = "white", linewidth = 0.1
  ) + # Draw only countries with billionaires
  geom_polygon(
    data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
    aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
    color = "white", linewidth = 0.1
  ) +
  scale_fill_viridis(
    option = "mako",
    name = "Number of\nbillionaires",
    trans = "log10", # Apply logarithmic transformation
    breaks = c(1, 5, 10, 50, 100, 500),
    direction = -1
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```

For this specific map, we found inspiration for the code among R resources[^map_tuto] found on the web and we then used AI assistance to help refine the R code and generate a clearer, more visual representation.


@fig-Distribution-of-billionaires illustrates the distribution of billionaires across various industries, highlighting that real estate is the most prominent sector for amassing such wealth, followed by retail and construction. A significant trend across almost all sectors is the clear dominance of self-made individuals over those who inherited their fortunes. This disparity is particularly pronounced in modern industries like technology, software, and hedge funds, where inherited wealth is virtually non-existent, whereas more traditional sectors like oil and banking show a relatively higher presence of generational wealth.


```{r}
#| label: fig-Distribution-of-billionaires
#| fig-cap: "Distribution of billionaires by sector in 2014"
#| echo: false
#| message: false
#| warning: false

Sectors <-
  Billionaires_2014 |>
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |>
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(
    cols = c(nb_self_made, nb_inherited),
    names_to = "origine",
    values_to = "count"
  ) |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  )


```

An analysis of billionaires' wealth between 1996 and 2014 reveals a profound structural change. Beyond the exponential growth in their net worth (notably a 261% increase since 2001), the most significant change lies in the source of their wealth: “self-made” entrepreneurs have now largely overtaken heirs.

Geographically, this wealth remains concentrated in high-income hubs (the United States, Western Europe, and East Asia), anchored in traditional sectors such as real estate, while increasingly driven by technology and finance. The emergence of this self-made and highly mobile elite suggests that their location depends less on tradition than on the economic environment. This brings us to our econometric analysis, in which we will examine how market size, openness, and taxation influence the spatial distribution of these global “titans.”

# First model : understdanding the spacial distribution of billionaires
## Initial study and early results

Our baseline regression, see @tbl-first-model, follows @eq-first_model : 

$$log(Nb) = \beta_0 + \beta_1 \cdot tax + \beta_2 \cdot log(GDP) + \beta_3 \cdot FDI + \beta_4 \cdot Capital$$ {#eq-first_model} 

It links the (log) number of billionaires ($Nb$) in a country to our taxation proxy ($tax$), economic size ($GDP$), openness ($FDI$), and investment capacity ($capital$). 

Overall, the model fits the data well (Adjusted $R^2= 0.770$; F-test p-value $< 10^{-14}$), suggesting these macroeconomic factors explain a large share of cross-country variation in billionaire presence.

Tax revenue enters positively ($\beta_1 = 0.022$) and is only marginally significant at the 10% level (p = 0.097). Interpreted as a proxy for overall fiscal pressure, this suggests that stronger tax capacity is associated with more, not fewer, billionaires. A plausible interpretation is that high-tax countries also tend to be more institutionally developed and able to provide high-quality public goods (education, health, infrastructure, regulatory stability) that support profitable business activity, so these indirect advantages may offset the direct cost of taxation. Accordingly, taxation does not automatically weaken a country’s ability to host top wealth holders, “competitiveness” can also rest on stability and market opportunities rather than low taxes alone.

Economic scale is the dominant factor: log GDP is large and highly significant, consistent with the idea that bigger markets generate more high-value firms, deeper capital markets, and more opportunities for large fortunes. FDI is also positive and significant, indicating that more open and internationally connected economies tend to host more billionaires. By contrast, capital formation is not statistically significant (p = 0.40), suggesting limited additional explanatory power once GDP and FDI are controlled for.

Overall, the baseline results indicate that higher tax revenue is not negatively associated with billionaire presence, while market size (GDP) and openness (FDI) are strongly associated with billionaire concentration.

```{r}
#| label: tbl-first-model
#| tbl-cap: "First model results"
#| message: false
#| warning: false
#| echo: false

# Data preparation

reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |>
    select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |>
    select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |>
    select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |>
    select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

# First model

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)

modelsummary(
  modele, 
  output = "kableExtra",
  fmt = 3,
  shape = term + statistic ~ model, 
  stars = TRUE,
  gof_map = c("r.squared", "adj.r.squared", "nobs")) |> 
  kable_styling(latex_options = c("striped", "scale_down", "hold_position"))

```


## Robustness check

Overall, the diagnostic tests support the validity of the OLS assumptions for our first model.
The VIF results[^multico] (around 1.2–1.3) indicate no multicollinearity, so the regressors are not strongly correlated.
The Shapiro–Wilk test[^shapiro] (p = 0.7815) suggests the residuals are approximately normally distributed.
The Breusch–Pagan test[^Breusch] (p = 0.7318) shows no evidence of heteroskedasticity, meaning the error variance can be treated as constant.
Taken together, these checks imply that our coefficient estimates and standard inference from this model are generally reliable.

The output tables for the different tests are presented in [Annex 1](#sec-one)

# Understanding and anticipating the behaviour of the wealthiest individuals

The previous baseline results describe where billionaires are most likely to be located at the country level, but they remain silent on how the individuals behind these fortunes may respond to policy changes. The next section therefore shifts from macro determinants of billionaire concentration to the micro-level question at the core of our study: understanding who is most likely to adjust their location and how we can anticipate mobility and tax-flight behaviour among the wealthiest individuals.
 
## Creation of a flight-risk score {#sec-ponderation}

We will therefore create in this section a flight risk score, based on four distinct component :

A tax burden score between 0 and 1 calculated according to the following standardization :

$$\frac{(\text{tax in country i})- (\text{min tax})}{(\text{max tax})- (\text{min tax})}$$ {#eq-standardization} 

This score is thus normalized between 0 (minimum taxation) and 1 (maximum taxation). We consider that higher taxation would encourage more tax evasion.  

Whether the billionaire's main activity belongs to an industry considered to be internationally mobile. Among the industries in the original dataset, we have identified the following as mobile : "Diversified financial",  "Technology-Computer", "Hedge funds", "Private equity/leverage buyout",      "Venture capital", "Media", "Money Management". Belonging to one of this industry gives a score of 1 and 0 otherwise as we believe that working in such a sector would facilitate mobility for the billionaire. 

The source of the billionaire's wealth, with a score of 1 if "self-made" and 0 if "inherited". This score is used to capture the idea that inherited wealth is more likely to be accompanied by a strong attachment to the territory of origin. The idea is that inherited wealth tends to involve real estate and historically established industries, which provide incentives to stay put. 

Proximity to a tax haven. Based on the official definition, we first drew up the following list of tax havens: "Switzerland", "Singapore", "Hong Kong", "Liechtenstein", "Monaco", "Luxembourg", "United Arab Emirates" and "Bermuda". We then identified the countries considered close enough to a tax haven to provide billionaires with a significant incentive to relocate. To do this, we took into account membership of the Schengen area for European countries (where mobility to the border of the tax haven is facilitated, whether for Switzerland, Luxembourg or Liechtenstein) and the sharing of a coastline with the same ocean for Asian and Caribbean tax havens. This gives us a score of 1 if close, 0 otherwise.

```{r}
#| label: Creation of the 4 scores
#| echo: false
#| message: false
#| warning: false


# Sectors considered "mobile"

mobile_sectors <- c("Diversified financial", "Technology-Computer", "Hedge funds", "Private equity/leverage buyout", "Venture capital", "Media", "Money Management")

# List of tax havens/refuges

tax_havens <- c(
  "Switzerland", "Singapore", "Hong Kong", "Liechtenstein",
  "Monaco", "Luxembourg", "United Arab Emirates", "Bermuda"
)

# Calcul of the 4 indicators

flight_risk_index <-
  Billionaires_2014 |>
  left_join(Tax_revenue_reduced |> select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  mutate(
    # tax burden score calculation
    tax_normalized =
      ifelse(tax != 0, (tax - min(tax, na.rm = TRUE)) / (max(tax, na.rm = TRUE) - min(tax, na.rm = TRUE)), 0),
    tax_normalized =
      ifelse(is.na(tax) == TRUE, 0, tax_normalized),
    # mobility score calculation
    mobile_sector = ifelse(industry %in% mobile_sectors, 1, 0),
    # source of wealth index computation
    selfmade = ifelse(is.na(selfmade) == TRUE, "self-made", selfmade),
    is_selfmade = ifelse(selfmade == "self-made", 1, 0),
    is_selfmade = replace_na(is_selfmade, 0),

    # proximity to tax haven index computation
    near_haven = case_when(
      COUNTRY %in% c(
        "France", "Germany", "Italy", "Spain", "United Kingdom",
        "Belgium", "Netherlands", "Austria", "Sweden", "Norway",
        "Portugal", "Denmark", "Czech Republic", "Finland", "Greece",
        "Poland", "Romania", "Cyprus", "Lithuania"
      ) ~ 1,
      COUNTRY %in% c(
        "China", "Japan", "South Korea", "India", "Malaysia",
        "Thailand", "Indonesia", "Taiwan", "Philippines", "Macau", "Vietnam"
      ) ~ 1,
      COUNTRY %in% c(
        "Saudi Arabia", "Kuwait", "Israel", "Turkey", "Egypt",
        "Lebanon", "Oman"
      ) ~ 1,
      COUNTRY %in% c(
        "United States", "Canada", "Brazil", "Mexico", "Colombia",
        "Venezuela", "St. Kitts and Nevis"
      ) ~ 1,
      TRUE ~ 0
    ),
    in_tax_haven = ifelse(COUNTRY %in% tax_havens, 1, 0)
  )
```

For each billionaire, four indicators are calculated, which are then combined to obtain the final flight risk score. For the calculation, we assigned the following coefficients :

$$
\text{Score} = 0.45 \cdot \text{Taxation} + 0.3 \cdot \text{Mobility} + 0.15 \cdot \text{Source of wealth} + 0.1 \cdot \text{Proximity}
$$ {#eq-weighting}

Taxation has the highest coefficient because it is theoretically the most significant incentive to leave. Mobility comes next, as it represents the most difficult physical constraint to overcome. The origin of wealth has a less significant coefficient because its impact remains limited due to the complexity of accurately establishing a billionaire's personal ties to their territory. The smallest coefficient is associated with proximity to a tax haven. Despite the physical constraint that distance represents, the economy remains globalized and billionaires have ample means to circumvent the issue.

Tax havens are assigned a default score of 0, and for the sake of clarity, categories are assigned according to the score: "Low" if below 0.3, “Medium” between 0.3 and 0.6, and "High" if above 0.6.

```{r}
#| label: Aggregation of the 4 scores
#| echo: false
#| message: false
#| warning: false

flight_risk_index <-
  flight_risk_index |>
  mutate(
    wealth_bil = networthusbillion,
    flight_risk_score = (tax_normalized * 0.45 +
      mobile_sector * 0.30 +
      is_selfmade * 0.15 +
      near_haven * 0.10),
    flight_risk_score = ifelse(in_tax_haven == 1, 0, flight_risk_score),
    risk_category = case_when(
      flight_risk_score < 0.3 ~ "Low",
      flight_risk_score >= 0.3 & flight_risk_score < 0.6 ~ "Medium",
      flight_risk_score >= 0.6 ~ "High"
    )
  )
```


## Conversion of the flight-score into a probability using a logistic transformation {#sec-logis}

*For this phase, we wanted to convert our scores into probabilities so that we could interpret them effectively. We based our work on Niculescu-Mizil and Caruana[^Niculescu] (2005) and found various online sources[^Sources] explaining the main principles of probability calculation using logistic regression. Finally, the use of AI enabled us to confirm our method and refine our code.*

We convert the flight-risk score into an expected flight probability using a logistic transformation. The score is a 0–1 index that ranks individuals by exposure to mobility incentives, but it cannot be read directly as a probability (or as the share of wealth that would relocate). The logistic function, or sigmoid function, converts a probability into a real score. Its initial form is as follows:


$$f(z) = \frac{1}{1+e^{-z}}$$ {#eq-logistic}

We therefore use its inverse, the logit function:

$$logit(p) = log(\frac{1}{1-p})$$ {#eq-logit}

This function can then be used to convert a probability into a score:

$$P_{prob} = \frac{1}{1+e^{-(\alpha + \beta S)}}$$ {#eq-prob_to_score}

Where $\beta$ represents the sensitivity of the leakage risk relative to the score and $\alpha$ is the adjustment constant (the intercept).

Because we do not observe actual relocation decisions in our data, we calibrate (rather than estimate with the Maximum Likelihood Method) the logistic parameters $(\alpha)$ and $(\beta)$ transparently using two anchor points from the score distribution : The score corresponding to the 10th percentile of billionaires $s_{low}$ (low-risk profiles) and the score corresponding to the 90th percentile $s_{high}$ (high-risk profiles). This "moderate mobility" calibration pins down $(\alpha)$ and $(\beta)$ uniquely and yields an average implied flight probability of about 10.06\% (as we draw it from @tbl-country-risk) in the sample. 

To justify our orders of magnitude (anchors $p_{10}=2\%$ and $p_{90}=25\%$ at horizon $T=15$ years), we start from a homogeneous benchmark directly grounded in empirical estimates of annual migration flows in the literature. Advani, Burgherr \& Summers (2025)[^Advani] estimate a semi-elasticity of emigration of about $-0.339$: a 1\% decrease in the net-of-average-tax rate increases the annual emigration rate by roughly 0.3 percentage points (temporarily). Applying this order of magnitude linearly to a 2\% shock (calibration approximation) yields an average annual increment $\mu \approx 0.66\%$ (i.e., $0.0066$). If all individuals responded identically (no heterogeneity), the implied 15-year cumulative probability would be

$$
p_{15} \;=\; 1 - (1-0.0066)^{15} \;\approx\; 9.4\%,
$$

which is very close to our calibrated mean (about 10.06\%)-hence our levels are consistent with empirically observed average responses. The remaining gap (10.06\% $>$ 9.4\%) is mechanically explained by the fact that our model introduces heterogeneity: we impose a right tail with highly mobile profiles (25\% at the 90th percentile) and strongly anchored profiles (2\% at the 10th percentile). This asymmetry pulls the mean upward and is consistent with documented stylized facts: Advani, Burgherr \& Summers also show that responsiveness is concentrated among individuals with weaker local economic ties (e.g., limited domestic-source income), who exhibit substantially higher migration elasticities than the average. 
Jakobsen et al.[^Jakobsen] similarly report an average semi-elasticity of out-migration of about $-0.17$, while noting that some subgroups (e.g., entrepreneurs) can display larger responses albeit less precisely estimated. 
In sum, our ``2\%--25\%'' calibration embeds heterogeneity that increases the weight of highly mobile profiles in the mean, while remaining anchored in empirically plausible magnitudes for migration flows and in the observed concentration of responses within a more sensitive minority.

These scores are then assigned two flight probabilities $p_{low}$ and $p_{high}$ in order to find the parameters after solving a system of two equations :

$$
\begin{cases} 
\beta = \frac{\text{logit}(p_{90}) - \text{logit}(p_{10})}{s_{high} - s_{low}} \\ 
\alpha = \text{logit}(p_{10}) - \beta \cdot s_{low} 
\end{cases}
$$ {#eq-system}

We can now use @eq-prob_to_score and convert our score into a probability $P_{prob}$.


```{r}
#| label: Conversion using the logistic function
#| echo: false
#| message: false
#| warning: false


logit <- function(p) log(p / (1 - p)) # Define the logit function: logit(p) = log(p / (1 - p))
# two anchor points from the empirical distribution of the score:
s_low <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.10, na.rm = TRUE)) #    s_low  = 10th percentile of flight_risk_score (low-risk profiles)
s_high <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.90, na.rm = TRUE)) #    s_high = 90th percentile of flight_risk_score (high-risk profiles)

# Target flight probabilities at these anchors (moderate mobility scenario):
# p_low  is the expected flight probability at s_low
# p_high is the expected flight probability at s_high
p_low <- 0.02 # ~2% expected flight at the low-risk (10th percentile) score
p_high <- 0.25 # ~25% expected flight at the high-risk (90th percentile) score

# Solve for beta and alpha in:
#    p(score) = plogis(alpha + beta * score)
# Using the logit transform:
#    logit(p_low)  = alpha + beta * s_low
#    logit(p_high) = alpha + beta * s_high
# Subtracting the two equations gives beta, then alpha follows.
beta <- (logit(p_high) - logit(p_low)) / (s_high - s_low)
alpha <- logit(p_low) - beta * s_low

# Compute the implied flight probability for each billionaire:

flight_risk_index <- 
  flight_risk_index |>
  mutate(
    flight_prob = ifelse(flight_risk_score!=0, 1/(1+exp(-(alpha + beta*flight_risk_score))), 0)
  )

```


## Analysis of the results

Now that we have converted the flight-risk score into a calibrated flight probability (flight_prob), we aggregate the results at the country level. This allows us to compare countries in terms of (i) billionaire concentration, (ii) total billionaire wealth, and (iii) both the average risk index and the average implied mobility (probability), focusing on countries with a sufficiently large billionaire sample ($>10$).

```{r}
#| label: tbl-country-risk
#| tbl-cap: "Top Country Risk Analysis"
#| echo: false
#| message: false
#| warning: false

country_risk <-
  flight_risk_index |>
  summarise(
    nb_billionaires = n(), # number of billionaires
    total_wealth = sum(wealth_bil, na.rm = TRUE), # total wealth ($bn), unit-corrected
    avg_flight_risk = mean(flight_risk_score, na.rm = TRUE), # average risk score (index)
    avg_flight_prob = mean(flight_prob, na.rm = TRUE), # average risk score (prob)
    nb_high_risk = sum(risk_category == "High", na.rm = TRUE), # average calibrated flight probability
    pct_high_risk = (nb_high_risk / n()) * 100, # % of high-risk individuals
    tax = mean(tax, na.rm = TRUE), # average tax proxy (same within a country)
    .by = COUNTRY
  ) |>
  arrange(desc(avg_flight_risk))

top_country_risk <-
  country_risk |>
  filter(nb_billionaires > 10) |>
  slice_max(order_by = avg_flight_prob, n = 10)

top_country_risk |> 
  select(1, 2, 5) |> 
  kable(format = "latex",
        booktabs = TRUE,
        longtable = TRUE,
        digits = 2) |> 
  kable_styling(latex_options = "hold_position", position = "center")

```

@tbl-country-risk aggregates the flight-risk indicators at the country level for countries with more than ten billionaires. It reports the average flight-risk score (our composite index) and the average implied flight probability (from the calibrated logistic mapping). 

The United Kingdom, Israel, and Sweden appear at the top with relatively high average probabilities, which fits an interpretation where a larger share of fortunes is connected to internationally mobile activities (finance, tech, media) and where cross-border relocation is comparatively frictionless. Malaysia and Thailand also rank relatively high despite smaller billionaire populations, suggesting a composition effect: although their numbers are smaller, the billionaire profiles present in these sectors may be more concentrated in less restrictive sectors, which drives up the average mobility indicator.

France sits in an intermediate position: its average flight probability is meaningful but below the top countries, consistent with the idea that part of French billionaire wealth is tied to established domestic firms and less mobile assets that anchor residence and business operations. The United States is a useful contrast. Despite having by far the largest number of billionaires, its average flight probability is only moderate. This supports a “home-market” mechanism: deep domestic opportunities and large market scale can reduce the incentive to relocate on average, even though there may still be a sizable high-mobility minority in absolute numbers.


```{r}
#| label: fig-Basic-analysis-plots
#| fig-cap: "Distribution of flight probability"
#| echo: false
#| message: false
#| warning: false

plot_distrib <- ggplot(flight_risk_index, aes(x = flight_prob))+
  geom_density()+
  labs(
    title = "Distribution of the flight probability",
    x = "Probability"
  )

plot_region <- ggplot(flight_risk_index, aes(x = flight_prob))+
  geom_histogram()+
  facet_wrap(~selfmade)+
  labs(
    title = "Distribution of the flight probability by source of wealth",
    x = "Probability"
  )

plot_region / plot_distrib

```

In the top panel, self-made billionaires display higher predicted mobility than inherited ones. Inherited wealth is concentrated at very low flight probabilities, consistent with fortunes that are more anchored in domestic assets and long-established family structures, which make relocation costlier. By contrast, the self-made distribution has a thicker right tail, indicating more individuals with medium-to-high flight probabilities. This fits an economic interpretation where self-made wealth is more often tied to internationally scalable businesses and financial assets that are easier to reorganize across borders.

The bottom density plot summarizes the overall distribution of flight probabilities in the sample. It is strongly right-skewed: most individuals cluster at low probabilities, while a smaller group forms a long tail toward higher values. This matters because it implies heterogeneity in behavioral responses. Indeed, the flight risk is not evenly spread across billionaires. The density also appears multi-peaked, which is consistent with the index being built from discrete components (mobile sector, proximity, self-made status) combined with a continuous tax term: different “profiles” of billionaires naturally bunch at different probability ranges.

Overall, @fig-Basic-analysis-plots suggest that mobility risk is not uniform across billionaires. Most individuals cluster at low predicted flight probabilities, especially among inherited fortunes, but there is a clear right tail driven largely by self-made profiles. In other words, the potential response to a wealth tax is concentrated in a smaller, more mobile subset rather than spread evenly across the entire billionaire population.




# Scenario testing : the impact of new taxation 

In this section, we move from descriptive evidence to policy evaluation. Building on our flight-risk framework, we simulate the expected fiscal impact of introducing a new 2% billionaire wealth tax under alternative institutional settings. The goal is to quantify how different degrees of international coordination affect wealth mobility, the erosion of the tax base, and ultimately the amount of revenue that can be effectively collected.

## Creating a scenario-testing function {#sec-creation}

We define a set of nested country coalitions to represent progressively broader degrees of international coordination. We start with the G7 and the G20, which capture coordination among major advanced economies and among the largest global economies, respectively. To extend the analysis beyond these institutional groups, we also build a broader “G20 + North” coalition by adding all countries classified as part of the Global North in our dataset, reflecting a wider agreement among high-income jurisdictions that host a large share of the taxable base. Finally, we consider a “global” benchmark that includes all countries in our sample, which serves as an upper bound on coordination. This tiered structure allows us to assess how expanding participation changes expected mobility incentives and the effective fiscal yield of the tax.

```{r}
#| label: Define G7 and G20 members
#| message: false
#| warning: false
#| echo: false

G7 <- c(
  "United States", "United Kingdom", "Germany", "France",
  "Italy", "Canada", "Japan"
)

G20 <- c(
  G7, "China", "India", "Brazil", "Russia", "Australia",
  "South Korea", "Mexico", "Indonesia", "Turkey", "Saudi Arabia",
  "Argentina", "South Africa"
)

NORTH <- flight_risk_index |>
  filter(north == 1) |>
  distinct(COUNTRY) |>
  pull(COUNTRY)

G20_North <- unique(c(G20, NORTH)) #Coalition 4 = G20 + North countries 

GLOBAL <- unique(flight_risk_index$COUNTRY) #All countries in our sample
```

We then implement a revenue simulation that converts our calibrated individual flight probability into fiscal outcomes under each policy setting. For each billionaire, the baseline flight probability $P_{prob}$ is scaled by a scenario-specific multiplier $\eta$ higher under unilateral action and lower under broader coordination-and capped at 1 to avoid probabilities above 100%, following @eq-Effective_prob, to give the effective flight probability $\mathcal{E}_p$ and its average $\mu_{\mathcal{E}}$:
 
$$\mathcal{E}_p = min\{ \frac{\eta}{1+e^{-(\alpha + \beta S)}} \ ; 1\}$$ {#eq-Effective_prob}

With a 2% tax rate on billionaire wealth and using initial wealth $\mathcal{I}$, the simulation computes the fled wealth $l$, remaining wealth (which will be taxed) $r$, potential revenue (if there is no leakage) $\mathcal{P}$ and actual revenue (after leakage) $A$. 

$$
\begin{cases} 
l =  \mathcal{E}_p \cdot \mathcal{I}\\ 
r =  \mathcal{I} - l \\
\mathcal{P} = \mathcal{I} \cdot 0.02 \\ 
A  = r \cdot 0.02
\end{cases}
$$ {#eq-simulation}

The difference between potential and actual revenue computed in @eq-simulation gives the revenue loss attributable to flight $\mathcal{L}$ of @eq-loss, and their ratio provides a collection-efficiency measure $CE$ of @eq-CE. More explicitly, it is defined as the share of the theoretical tax yield that is effectively collected once flight is taken into account. It serves as a summary indicator of tax-base erosion: higher values imply limited leakage and stronger enforcability of the policy, whereas lower values indicate substantial revenue losses due to mobility.

$$\mathcal{L} = \mathcal{P} - A$$ {#eq-loss}

$$CE = \sum_{i=1}^{n}(\frac{A_i}{\mathcal{P}_i})\cdot 100$$ {#eq-CE}



```{r}
#| label: Tax revenue simulation function
#| echo: false
#| message: false
#| warning: false


simulate_tax_scenario <- function(data, scenario_name, flight_multiplier) {

  results <- data |>
    mutate(
      effective_flight_prob = flight_prob * flight_multiplier, 
      effective_flight_prob = pmin(effective_flight_prob, 1), 
      initial_wealth = wealth_bil,
      fled_wealth = initial_wealth * effective_flight_prob, 
      taxable_wealth = initial_wealth - fled_wealth, 
      potential_revenue = initial_wealth * 0.02, 
      actual_revenue = taxable_wealth * 0.02, 
      lost_revenue = potential_revenue - actual_revenue, 
    )


  # Global summary
  global_summary <- results |>
    summarise(
      scenario = scenario_name,
      total_billionaires = n(),
      avg_flight_prob = mean(effective_flight_prob) * 100,
      total_wealth_billion = sum(initial_wealth),
      fled_wealth_billion = sum(fled_wealth),
      potential_revenue_billion = sum(potential_revenue),
      actual_revenue_billion = sum(actual_revenue),
      lost_revenue_billion = sum(lost_revenue),
      collection_efficiency = (actual_revenue_billion / potential_revenue_billion) * 100
    )

  # Summary by country (top contributors)
  country_summary <- results |>
    group_by(COUNTRY) |>
    summarise(
      scenario = scenario_name,
      nb_bill = n(),
      potential = sum(potential_revenue),
      actual = sum(actual_revenue),
      lost = sum(lost_revenue)
    ) |>
    arrange(desc(actual))

  return(list(global = global_summary, by_country = country_summary))
}
```

## Application on different scenarios

Finally, we compare a unilateral implementation (France alone) to increasingly broad coalitions (G7, G20, G20+North, and a quasi-global benchmark). As explained in @sec-creation, each scenario is characterized by a specific flight multiplier $\eta$ that captures how coordination constrains relocation opportunities: the multiplier is highest under unilateral action and progressively lower as the coalition expands. 

Our flight multipliers are calibrated on the basis of recent empirical studies of tax mobility. The unilateral multiplier ($\eta_{FR}$ = 1.5) relies on the elasticity of 7.96 estimated by Agrawal et al.\ (2023) in the Spanish context, where a full wealth-taxation differential generated a 7.5\% increase in the wealthy population in the zero-rate region over 6 years. Our unilateral French scenario, with a 2\% rate, generates an expected average leakage of 15.1\%, consistent with this elasticity and with the historical experience of the ISF.
The decreasing multipliers for the coordinated scenarios ($\eta_{G7}$ = 0.9 for the G7, $\eta_{G20}$= 0.7 for the G20, $\eta_{G20+}$ = 0.55 for G20+North, $\eta_{all}$ = 0.40 for quasi-global) reflect the progressive reduction in the alternative destinations available. This approach is consistent with the conclusions of Jakobsen et al.\ (2024) showing ``modest'' migratory responses in a Scandinavian coordination context, the analysis of Zucman (2024) demonstrating a collection rate of 95.9\% under global coordination versus 80\% under unilateral implementation, and the established fact that only 9\% of global billionaires reside outside their country of citizenship (Forbes, in Zucman 2024).


As a robustness check, the same simulations could be rerun using alternative flight multipliers to assess how sensitive our results are to the assumed strength of behavioral responses. In practice, we can define “low-mobility” and “high-mobility” variants around the baseline ladder (slightly higher or lower $\eta$ in each scenario) and verify whether the qualitative ranking of scenarios and the main conclusions on revenue losses and collection efficiency remain unchanged.

```{r}
#| label: tbl-scenarios
#| tbl-cap: "Tax scenarios resume"
#| echo: false
#| message: false
#| warning: false

sc_fr <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY == "France"), "France alone", 1.5)
sc_g7 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G7), "G7 Coordination", 0.9)
sc_g20 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G20), "G20 Coordination", 0.7)
sc_g20_north <- simulate_tax_scenario(
  flight_risk_index |> filter(COUNTRY %in% G20_North),
  "Expanded coordination (G20 + North)",
  0.55
)

sc_global <- simulate_tax_scenario(
  flight_risk_index |> filter(COUNTRY %in% GLOBAL),
  "Near-global coordination",
  0.40
)

summary_table_all_display <- bind_rows(sc_fr$global, sc_g7$global, sc_g20$global, sc_g20_north$global, sc_global$global) |>
  transmute(
    Scenario = scenario,
    "n" = total_billionaires,
    "$\\mu_{\\mathcal{E}}$ (%)" = round(avg_flight_prob, 1),
    "$\\mathcal{P}$ ($bn)" = round(potential_revenue_billion, 3),
    "A ($bn)" = round(actual_revenue_billion, 3),
    "$\\mathcal{L}$ ($bn)" = round(lost_revenue_billion, 3),
    "CE (%)" = round(collection_efficiency, 1)
  )

summary_table_all <- bind_rows(sc_fr$global, sc_g7$global, sc_g20$global, sc_g20_north$global, sc_global$global) |>
  transmute(
    Scenario = scenario,
    `# billionaires` = total_billionaires,
    `Expected flight rate (%)` = round(avg_flight_prob, 1),
    `Potential revenue ($bn)` = round(potential_revenue_billion, 3),
    `Actual revenue ($bn)` = round(actual_revenue_billion, 3),
    `Lost revenue ($bn)` = round(lost_revenue_billion, 3),
    `Collection efficiency (%)` = round(collection_efficiency, 1)
  )

kable(
  summary_table_all,
  format = "latex", 
  booktabs = TRUE,
  digits = 2,
  escape = FALSE,
  col.names = c(
    "Scenario",
    "n",
    "$\\mu_{\\mathcal{E}}$ (\\%)",
    "$\\mathcal{P}$",
    "A" ,
    "$\\mathcal{L}$",
    "CE (\\%)"
  )
) |>
kable_styling(latex_options = "hold_position", position = "center", full_width = FALSE)|>
  footnote(
  general = "n : number of billionaires ; $\\\\mu_{\\\\mathcal{E}}$ : average flight probability ; $\\\\mathcal{P}$ : potential revenue ; A : actual revenue ; $\\\\mathcal{L}$ : lost revenue ; CE : collection efficiency ; values in bn USD",
  general_title = "Note: ",
  footnote_as_chunk = TRUE,
  threeparttable = TRUE,
  escape = FALSE,
  fixed_small_size = TRUE
)
```


In Zucman (2024)[^Zucman_paper], the notion of "collection efficiency" (effective collection rate) can be read as the ratio between revenues actually collected and potential revenues under perfect enforcement of the statutory rate. Zucman indicates that a 2\% minimum tax on billionaire wealth would generate \$242bn under perfect enforcement, and \$193bn when assuming that 20\% of potential revenues are lost to avoidance/evasion. This immediately implies an implicit collection efficiency:

$$
CE^{Zucman}=\frac{193}{242}\approx 0.80,
$$ 

i.e.an aggregate loss of $1-0.80=0.20$, namely 20\%.

Our `"France alone" scenario (@tbl-scenarios) computes exactly the same quantity, derived from @eq-CE as:

$$
CE=\frac{\text{Actual revenue}}{\text{Potential revenue}}.
$$ {#eq-derived}

The table reports potential revenue $P=3.66$ (in \$bn) and realized revenue $A=28.7$, which implies:

$$
CE^{FR}=\frac{28,7}{36,6}\approx 0.784,
$$
i.e. 78.4\%, and therefore an associated loss rate of $1-0.784=0.216$, i.e. 21.6\%. In other words, our unilateral simulation produces a collection efficiency very close to the "low-enforcement" bound illustrated by Zucman[^Zucman_paper] (a difference of 1.6 percentage points).

This proximity is substantive because it is obtained under structurally different assumptions. In Zucman's paper, the 20\% loss is introduced in a top-down manner as a uniform haircut applied to potential revenues. In our framework, flight is endogenous at the micro level : it results from a distribution of individual flight probabilities derived from a composite score that explicitly incorporates heterogeneity (wealth origin, sectoral mobility, proximity to low-tax jurisdictions, and fiscal pressure), and then calibrated through a logistic transformation. The fact that we obtain almost the same collection efficiency as Zucman despite this heterogeneity is informative: it suggests that the aggregate assumption of "about 20\% losses" can emerge naturally from a bottom-up model in which mobility is concentrated in a high-risk minority (and, in our data, in which wealth is more concentrated among these profiles), rather than being imposed uniformly.


```{r}
#| message: false
#| warning: false
#| echo: false
#| label: fig-coordination
#| fig-cap: "Tax-base erosion declines with stronger coordination"

loss_plot_df <- summary_table_all |>
  mutate(
    Scenario = factor(
      Scenario,
      levels = c("France alone", "G7 Coordination", "G20 Coordination",
                 "Expanded coordination (G20 + North)", "Near-global coordination")
    ),
    # useful derived metric to interpret shape independent of scale
    `Loss rate (%)` = 100 * (`Lost revenue ($bn)` / `Potential revenue ($bn)`)
  )

ggplot(loss_plot_df, aes(x = Scenario, y = `Loss rate (%)`, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    x = NULL,
    y = "Lost revenue as % of potential revenue"
  ) 
```

@fig-coordination, consistent with @tbl-scenarios, highlights the decisive value of international coordination in making a tax on billionaires truly effective. As one moves from a unilateral implementation (France alone) to broader coalitions (G7, G20, G20+North, quasi-global), tax-base erosion decreases in a clear and monotonic way: the relative loss rate falls from 18.7\% to 4.0\%. This trajectory directly captures the central economic mechanism of coordination in an open economy: the broader the agreement, the more it reduces the set of credible fallback jurisdictions ("outside options") and thus the elasticity of the tax base to the reform. This result is precisely in line with the argument of Zucman[^Zucman_paper], according to which a minimum tax on billionaires becomes fully applicable only if the exit option is neutralized through international coordination. In our simulations, this progressive compression of outside options translates mechanically into a continuous increase in collection efficiency (from 78.4\% to 95.9\%): a growing share of the theoretical yield becomes effectively collectible. In other words, even if the expansion of the coalition also increases the taxable base and potential revenues, the key message is behavioral and institutional: coordination reduces flight and transforms a fragile tax into a robust fiscal instrument, once participation becomes sufficiently broad.



# Limitations

Several limitations should be noted with regard to this study. 

The first concerns the data used. The dataset on which our analysis is based stops at 2014, omitting potential changes in government debt dynamics and the wealth of billionaires over the last decade. The COVID crisis, for example, introduced an unprecedented dynamic with an explosion in the wealth of the top 0.01% while the global economy stagnated[^stagnated]. Updating the database would therefore enable more accurate results and analyses, reflecting the latest global economic trends. Likewise, since data on taxation by country is particularly tedious to collect and standardize, we chose to use the share of tax revenues in GDP as a proxy. Developing a better indicator by obtaining the rates applied at source would be a way to further improve the accuracy of our study.

From an econometric standpoint, while our first model accounts for GDP, FDI, and capital formation, it does not explicitly control for political stability, judicial independence, or the quality of life index. These omitted variables are likely correlated with both high tax revenues and billionaire presence, potentially leading to an overestimation of the positive relationship between taxation and wealth concentration. The diagnostic tests were performed on a sample of 52 observations, primarily representing countries with a significant billionaire presence. This small sample size may limit the generalization of the results to smaller or developing economies where the trade-off between taxation and competitiveness might follow a different trajectory.

Regarding the creation of our risk index, our mobility score is binary (1 or 0). This may oversimplify reality: a billionaire in the tech industry may be very locally rooted due to their ecosystem of engineers, while a billionaire in real estate may own highly liquid international assets. A more refined calculation of this score would improve the accuracy of our results.

Throughout our study, we repeatedly made calibration choices. In @sec-ponderation, we justified our weighting choices for calculating the leakage risk score based on economic intuition. In a future extension of the study, it would be relevant to test other scenarios in order to evaluate the resulting outcomes. In addition, we calibrated in @sec-logis our parameters by extending and adapting approaches developed in other research articles. It would be worthwhile to test other calibrations to confirm the robustness of our model and explore other economic scenarios. 

Following on from this article, a more detailed and individual analysis of billionaires' behavior could be conducted. When we attempted to calculate the probability of French billionaires leaving the country in scenario 1, see [Annex 2](#sec-two), certain insights emerged that should be explored further in future work.  

Finally, our model treats the billionaire population as static; it does not account for the dynamic impact of a 2% annual tax on capital accumulation over time, which could reduce the long-term growth rate of the tax base itself, even in the absence of physical flight.

# Conclusion

  This study examines the potential impact of a tax on billionaire fortunes on the spatial distribution of wealth and on mobility behaviors, using data covering 1,652 billionaires across 52 countries in 2014. We articulate two complementary components: on the one hand, a cross-sectional analysis of the macroeconomic determinants of the concentration of billionaires (aggregate tax pressure, market size, openness); on the other hand, a micro-founded approach aimed at anticipating tax flight, based on the construction of an individual risk score (sectoral mobility, origin of wealth, proximity to low-tax jurisdictions, tax pressure) and its conversion into relocation probabilities via a logistic transformation calibrated on empirical orders of magnitude from the literature on ultra-rich migration.

Our results reveal three main lessons. First, aggregate tax pressure (approximated by tax revenues as a percentage of GDP) does not exert a significant deterrent effect on the presence of billionaires: jurisdictions with high fiscal capacity offset this apparent cost through attractive non-tax factors (amenities, public goods and services, agglomeration effects), which dominate international comparisons and can mask the effect of taxes on location choices (Kleven et al., 2020). Second, economic size and international openness largely dominate the equation: a doubling of GDP is associated with an 89.5\% increase in the number of billionaires. The overall pattern is consistent with market scale effects highlighted by Neumayer (2004). Third, heterogeneity in mobility profiles is pronounced: our calibrated logistic model predicts an average probability of flight of 10.06\% over a fifteen-year horizon. This average nevertheless masks a strong concentration of risk among “self-made” entrepreneurs operating in dematerialized sectors (technology, finance), whose probability reaches 25% at the 90th percentile (versus only 2\% at the 10th percentile for heirs anchored in real estate or traditional industry).

Beyond these descriptive findings, this paper contributes to the literature on the tax elasticity of the ultra-rich by proposing a methodology for ex ante quantification of flight risk. This approach explicitly incorporates sectoral and wealth heterogeneity among the ultra-rich, whereas most studies focus on average aggregated elasticities.

Our scenario simulations show that a 2\% tax applied unilaterally by France would lead to a revenue loss of 18.7\% due to relocations, whereas international coordination (G20 or quasi-global) would bring this flight rate down to 4.1\%, raising collection effectiveness to 95.9\%.

The strong concentration of flight risk among a mobile minority indicates that the exit channel remains active and asymmetric: the most elastic profiles (those whose wealth is least territorially anchored) exert a disproportionate leverage effect on potential fiscal losses.

The final simulation confirms that multilateral coordination radically transforms the equation by mechanically reducing exit options, validating the theoretical hypothesis that taxing the ultra-rich requires collective action to be fiscally sustainable (Zucman, 2024).

In sum, our results suggest that the budgetary feasibility of a tax on billionaires depends less on the nominal level of the rate than on the institutional architecture of its implementation: without international coordination, the risk of a fiscal “race to the bottom” remains high for jurisdictions acting alone.

\newpage

# Annex

Given the word limit, we were unable to explore all results in greater depth. The figures presented in the appendix are provided for complementary purposes and left to the reader’s judgment, who may draw their own conclusions from them.

## Robustness check tests {#sec-one}

```{r}
#| label: tbl-vif
#| tbl-cap: "Multicolinearity analysis (VIF)"
#| message: false
#| warning: false
#| echo: false

library(car)

vif_df <- data.frame(
  Variable = names(vif(modele)),
  VIF = as.numeric(vif(modele))
)

kable(vif_df,
      col.names = c("Independent Variable", "VIF"),
      digits = 3,
      booktabs = TRUE) |> 
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```


```{r}
#| label: tbl-shapiro
#| tbl-cap: "Shapiro-Wilk"
#| message: false
#| warning: false
#| echo: false

shapiro_test <- tidy(shapiro.test(residuals(modele)))

kable(shapiro_test,
      col.names = c("W", "p-value", "Méthode"),
      digits = 4,
      booktabs = TRUE) |> 
  kable_styling(latex_options = "hold_position")
```


```{r}
#| label: tbl-breusch-pagan
#| tbl-cap: "Breusch-Pagan"
#| message: false
#| warning: false
#| echo: false


bp <- bptest(modele)

bp_df <- data.frame(
  Test = "Studentized Breusch-Pagan",
  Test_statistic = as.numeric(bp$statistic),
  p_value = as.numeric(bp$p.value)
)

kable(bp_df[, 2:3], 
      digits = 4, 
      booktabs = TRUE) |> 
  kable_styling(latex_options = "hold_position", 
                font_size = 9)

```


## Top contributors to revenue loss among french billionaires {#sec-two}

```{r}
#| label: tbl-Top-5-contributors-loss
#| tbl-cap: "Top 5 contributors to revenue loss"
#| message: false
#| warning: false
#| echo: false
top_loss_fr <- flight_risk_index |>
  filter(COUNTRY == "France") |>
  mutate(
    p_france = pmin(flight_prob * 1.5, 1),
    potential_rev = wealth_bil * 0.02,
    actual_rev = wealth_bil * (1 - p_france) * 0.02,
    lost_rev = potential_rev - actual_rev
  ) |>
  arrange(desc(lost_rev)) |>
  select(name, wealth_bil, p_france, lost_rev, industry, selfmade) |>
  head(10)

kable(top_loss_fr, 
             booktabs = TRUE, 
             digits = 2) |> 
  kable_styling(latex_options = "hold_position")
```


\newpage

# Bibliography

[^Zucman_paper]: See *A blueprint for a coordinated minimum effective taxation standard for ultra-high-net-worth individuals*, @Zucman2024 

[^World_inequality]: From @WorldInequality2026

[^French_debt]: See *5 minutes pour comprendre la dette publique*, @FrenchDebt  

[^map_tuto]: Helpful resources found at *How to create a map using ggplot2: The best reference*, @map_tuto

[^shapiro]: Based on *Normality test in R*, @shapiro 

[^multico]: Based on *Multicollinearity essentials and VIF in R*, @multico 

[^Breusch]: Based on *How to perform a Breusch-Pagan test in R*, @Breusch 

[^Niculescu]: See *Predicting good probabilities with supervised learning*, @Niculescu2005

[^Sources]: From [Probability Calculation Using Logistic Regression](https://docs.tibco.com/pub/sfire-dsc/6.5.0/doc/html/TIB_sfire-dsc_user-guide/GUID-C4D05ED0-3392-4407-B62A-7D29B26DC566.html) and [Logistic Regression – Explained](https://towardsdatascience.com/logistic-regression-explained-593e9ddb7c6c/)

[^Advani]: See *Taxation and migration by the super-rich*, @Advani2025 

[^Jakobsen]: See *Taxing top wealth: Migration responses and their aggregate economic implications*, @Jakobsen2024

[^stagnated]: From *World Inequality Report 2022*, @WIR2022

[^IMF]: Data from [@IMFDataExplorer]