---
title: "Final_phase"
format: html
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr)


Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv",
  skip = 3
)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv",
  skip = 3
)

Capital_formation <- read_csv("Database_project/Capital_formation.csv",
  skip = 3
)

Population <- read_csv("Database_project/Population.csv",
  skip = 3
)


useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |>  # standardization of names
  rename( COUNTRY = citizenship)




Billionaires_1996 <- 
  Billionaires |> 
  filter(year == "1996")

Billionaires_2001 <- 
  Billionaires |> 
  filter(year == "2001")

Billionaires_2014 <- 
  Billionaires |> 
  filter(year == "2014")




Billionaires_wide <- 
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Creating a list of every country represented in the dataset
Country <-
  Billionaires |>
  distinct(COUNTRY)

# Creating by hand, with the help of AI, a text list of those countries

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- c(
  "Bahrain, Kingdom of" = "Bahrain",
  "China, People's Republic of" = "China",
  "Egypt, Arab Republic of" = "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" = "Hong Kong",
  "Kazakhstan, Republic of" = "Kazakhstan",
  "Liechtenstein, Principality of" = "Liechtenstein",
  "Lithuania, Republic of" = "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" = "Macau",
  "Netherlands, The" = "Netherlands",
  "Poland, Republic of" = "Poland",
  "Russian Federation" = "Russia",
  "Korea, Republic of" = "South Korea",
  "Eswatini, Kingdom of" = "Swaziland",
  "Taiwan Province of China" = "Taiwan",
  "Tanzania, United Republic of" = "Tanzania",
  "Türkiye, Republic of" = "Turkey",
  "Venezuela, República Bolivariana de" = "Venezuela"
)

IMF <-
  IMF |>
  mutate(COUNTRY = recode(COUNTRY, !!!map_IMF))

# Selecting only the countries also present in the core dataset

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)


WB_names <-
  FDI |>
  distinct(`Country Name`)



map_WB <- c(
  "Czechia" = "Czech Republic",
  "Egypt, Arab Rep." = "Egypt",
  "Hong Kong SAR, China" = "Hong Kong",
  "Macao SAR, China" = "Macau",
  "Russian Federation" = "Russia",
  "Korea, Rep." = "South Korea",
  "Venezuela, RB" = "Venezuela",
  "Viet Nam" = "Vietnam"
)


# Using the map object creatd previously to standardize country names

FDI <-
  FDI |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object created previously to standardize country names

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object creatd previously to standardize country names

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object created previously to standardize country names

Population <-
  Population |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)
```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) PORTRAIT GLOBAL DES MILLIARDAIRES

### 1.1 Cartographie mondiale des milliardaires (2014)

```{r}
#| message: false
#| warning: false

library(maps)
library(viridis)
library(ggrepel)
library(broom)
library(scales)
library(patchwork)

# Agrégation par pays : nombre et richesse totale
billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(
    nb_billionaires = n(),
    total_wealth = sum(networthusbillion, na.rm = TRUE),
    avg_wealth = mean(networthusbillion, na.rm = TRUE)
  ) |>
  ungroup()

# Préparation des données cartographiques
world_map <- map_data("world")

# Mapping des noms de pays pour correspondance avec map_data
country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), 
                         country_mapping[COUNTRY], 
                         COUNTRY))

# Jointure avec la carte
world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region")

# Carte : Nombre de milliardaires par pays
map_count <- ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group),
               fill = "grey90", color = "white", linewidth = 0.1) +
  geom_polygon(data = world_billionaires |> filter(!is.na(nb_billionaires)), 
               aes(x = long, y = lat, group = group, fill = nb_billionaires),
               color = "white", linewidth = 0.1) +
  scale_fill_viridis(option = "plasma", 
                     name = "Nombre de\nmilliardaires",
                     trans = "log10",
                     breaks = c(1, 5, 10, 50, 100, 500))+
  theme_void() +
  labs(title = "Cartographie mondiale des milliardaires (2014)",
       subtitle = "Distribution géographique des ultra-riches") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```

### 1.3 Self-made vs Héritiers
```{r}

```


### 1.4 Répartition par secteur - Top 10
```{r}
secteurs <- Billionaires_2014 |>
  filter(!is.na(sector)) |>
  group_by(sector) |>
  summarise(n = n(), richesse = sum(networthusbillion, na.rm = TRUE)) |>
  arrange(desc(richesse)) |>
  head(10)

ggplot(secteurs, aes(x = reorder(sector, richesse), y = richesse)) +
  geom_col(fill = "purple") +
  coord_flip() +
  labs(title = "Top 10 secteurs par richesse totale",
       x = "", y = "Richesse (milliards $)") +
  theme_minimal()
```

## 2) ARBITRAGE FISCAL ET GÉOGRAPHIQUE

## 3) RÉGRESSION MULTIPLE : QUELS DÉTERMINANTS ?
```{r}
reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |> 
              select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |> select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |> select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |> select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(log_bill = log(nb_bill + 1),log_gdp = log(gdp)) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)
summary(modele)
```
