---
title: "Final_phase"
format: pdf
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr)
library(maps)
library(viridis)
library(ggrepel)
library(broom)
library(scales)
library(patchwork)
library(ggthemes)

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv",
  skip = 3
)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv",
  skip = 3
)

Capital_formation <- read_csv("Database_project/Capital_formation.csv",
  skip = 3
)

Population <- read_csv("Database_project/Population.csv",
  skip = 3
)


useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |>  # standardization of names
  rename( COUNTRY = citizenship)




Billionaires_1996 <- 
  Billionaires |> 
  filter(year == "1996")

Billionaires_2001 <- 
  Billionaires |> 
  filter(year == "2001")

Billionaires_2014 <- 
  Billionaires |> 
  filter(year == "2014")




Billionaires_wide <- 
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Creating a list of every country represented in the dataset
Country <-
  Billionaires |>
  distinct(COUNTRY)

# Creating by hand, with the help of AI, a text list of those countries

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- c(
  "Bahrain, Kingdom of" = "Bahrain",
  "China, People's Republic of" = "China",
  "Egypt, Arab Republic of" = "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" = "Hong Kong",
  "Kazakhstan, Republic of" = "Kazakhstan",
  "Liechtenstein, Principality of" = "Liechtenstein",
  "Lithuania, Republic of" = "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" = "Macau",
  "Netherlands, The" = "Netherlands",
  "Poland, Republic of" = "Poland",
  "Russian Federation" = "Russia",
  "Korea, Republic of" = "South Korea",
  "Eswatini, Kingdom of" = "Swaziland",
  "Taiwan Province of China" = "Taiwan",
  "Tanzania, United Republic of" = "Tanzania",
  "Türkiye, Republic of" = "Turkey",
  "Venezuela, República Bolivariana de" = "Venezuela"
)

IMF <-
  IMF |>
  mutate(COUNTRY = recode(COUNTRY, !!!map_IMF))

# Selecting only the countries also present in the core dataset

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)


WB_names <-
  FDI |>
  distinct(`Country Name`)



map_WB <- c(
  "Czechia" = "Czech Republic",
  "Egypt, Arab Rep." = "Egypt",
  "Hong Kong SAR, China" = "Hong Kong",
  "Macao SAR, China" = "Macau",
  "Russian Federation" = "Russia",
  "Korea, Rep." = "South Korea",
  "Venezuela, RB" = "Venezuela",
  "Viet Nam" = "Vietnam"
)


# Using the map object creatd previously to standardize country names

FDI <-
  FDI |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object created previously to standardize country names

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object creatd previously to standardize country names

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Using the map object created previously to standardize country names

Population <-
  Population |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

# Selecting only the countries also present in the core dataset

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

# Creating a sub dataset for future studies with only the years
# also present in the core dataset

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)
```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) State of the world of billionaires in 2014

### 1.1 Cartographie mondiale des milliardaires

```{r}
#| message: false
#| warning: false

# Aggregation by country: number and total wealth
billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data
world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format
country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |> 
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
                         country_mapping[COUNTRY], # Apply mapping if found
                         COUNTRY)) # Keep original name otherwise

# Join billionaire data with map data
world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country
map_count <- ggplot() +
  geom_polygon(data = world_map, # Draw all countries as base layer
               aes(x = long, y = lat, group = group), # Define coordinates and grouping
               fill = "grey90", color = "white", linewidth = 0.1) + # Draw only countries with billionaires
  geom_polygon(data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
               aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
               color = "white", linewidth = 0.1) +
  scale_fill_viridis(option = "plasma", 
                     name = "Number of\nbillionaires",
                     trans = "log10", # Apply logarithmic transformation
                     breaks = c(1, 5, 10, 50, 100, 500),
                     direction=-1)+
  theme_void() + # Remove axes and background for clean map
  labs(title = "Global distribution of billionaires in 2014")+
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```
For this specific map, we used AI assistance to help refine the R code and generate a clearer, more visual representation. We already had the initial intuition and the base approach, but we chose this mapping option because it is easier to read and more informative than a standard “classic” ggplot output.

This map shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs.


### 1.2 Caracteristics and origin of billionaires' wealth 

```{r}
#| message: false
#| warning: false

Sectors <-
  Billionaires_2014 |> 
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |> 
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(cols = c(nb_self_made, nb_inherited), 
               names_to = "origine", 
               values_to = "count") |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Distribution of billionaires by sector and origin of wealth in 2014",
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  ) +
  theme_clean()

```

```{r}
#| message: false
#| warning: false

Source <- 
  Billionaires |> 
  drop_na(selfmade) |> 
  summarise(n = n(), .by = c(selfmade, year)) 

ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(title = "Répartition de la richesse par année",
       y = "Année",
       x = "Nombre",
       fill = "Origine") +
  theme_clean()
```

## 2) ARBITRAGE FISCAL ET GÉOGRAPHIQUE

## 3) RÉGRESSION MULTIPLE : QUELS DÉTERMINANTS ?
```{r}
# =============================================================================
# RÉGRESSION DE BASE + TESTS APPROFONDIS
# =============================================================================

## Préparation des données (régression originale)
reg_data <- Billionaires_2014 %>%
  group_by(COUNTRY) %>%
  summarise(nb_bill = n()) %>%
  left_join(Tax_revenue_reduced %>% select(COUNTRY, tax = `2014`), by = "COUNTRY") %>%
  left_join(IMF_reduced %>% select(COUNTRY, gdp = `2014`), by = "COUNTRY") %>%
  left_join(FDI_reduced %>% select(COUNTRY, fdi = `2014`), by = "COUNTRY") %>%
  left_join(Capital_formation_reduced %>% select(COUNTRY, capital = `2014`), by = "COUNTRY") %>%
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) %>%
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

## MODÈLE DE BASE
modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)

cat("=== RÉGRESSION DE BASE ===\n")
summary(modele)

cat("\n=== INTERPRÉTATION DES COEFFICIENTS ===\n")
cat("Tax coefficient:", round(coef(modele)["tax"], 4), "\n")
cat("P-value:", round(summary(modele)$coefficients["tax", "Pr(>|t|)"], 4), "\n")
cat("Interprétation: Une hausse de 1% de taxation (en % du PIB) est associée à une variation de",
    round(coef(modele)["tax"], 3), "du log du nombre de milliardaires\n")
cat("Soit environ", round((exp(coef(modele)["tax"]) - 1) * 100, 2), 
    "% de variation du nombre de milliardaires.\n\n")


# =============================================================================
# TEST 1 : ANALYSE DE LA MULTICOLINÉARITÉ (VIF)
# =============================================================================

cat("\n=== TEST 1 : VARIANCE INFLATION FACTOR (VIF) ===\n")
cat("Vérifie si les variables sont trop corrélées entre elles\n\n")

# Calcul manuel du VIF
vif_values <- sapply(c("tax", "log_gdp", "fdi", "capital"), function(var) {
  formula <- as.formula(paste(var, "~ tax + log_gdp + fdi + capital -", var))
  r2 <- summary(lm(formula, data = reg_data))$r.squared
  1 / (1 - r2)
})

vif_table <- data.frame(
  Variable = c("tax", "log_gdp", "fdi", "capital"),
  VIF = vif_values
)
print(vif_table)

cat("\nInterprétation VIF :\n")
cat("- VIF < 5 : Pas de problème\n")
cat("- VIF 5-10 : Multicolinéarité modérée\n")
cat("- VIF > 10 : Multicolinéarité problématique\n\n")


# =============================================================================
# TEST 2 : TEST DE NORMALITÉ DES RÉSIDUS (Shapiro-Wilk)
# =============================================================================

cat("\n=== TEST 2 : NORMALITÉ DES RÉSIDUS ===\n")
shapiro_test <- shapiro.test(residuals(modele))
cat("Shapiro-Wilk test:\n")
cat("W =", round(shapiro_test$statistic, 4), "\n")
cat("P-value =", round(shapiro_test$p.value, 4), "\n")
cat("Interprétation:", ifelse(shapiro_test$p.value > 0.05, 
                              "Résidus normalement distribués ✓", 
                              "Résidus non normaux ✗"), "\n\n")


# =============================================================================
# TEST 3 : TEST D'HÉTÉROSCÉDASTICITÉ (Breusch-Pagan)
# =============================================================================

cat("\n=== TEST 3 : HÉTÉROSCÉDASTICITÉ (Breusch-Pagan) ===\n")

# Test manuel de Breusch-Pagan
residuals_sq <- residuals(modele)^2
bp_model <- lm(residuals_sq ~ tax + log_gdp + fdi + capital, data = reg_data)
bp_stat <- nobs(modele) * summary(bp_model)$r.squared
bp_pvalue <- pchisq(bp_stat, df = 4, lower.tail = FALSE)

cat("BP statistic =", round(bp_stat, 4), "\n")
cat("P-value =", round(bp_pvalue, 4), "\n")
cat("Interprétation:", ifelse(bp_pvalue > 0.05, 
                              "Homoscédasticité (variance constante) ✓", 
                              "Hétéroscédasticité détectée ✗"), "\n\n")


# =============================================================================
# TEST 4 : INFLUENCE DES OBSERVATIONS (Cook's Distance)
# =============================================================================

cat("\n=== TEST 4 : OBSERVATIONS INFLUENTES ===\n")

cooks_d <- cooks.distance(modele)
influential <- which(cooks_d > 4/nobs(modele))

cat("Seuil Cook's distance:", round(4/nobs(modele), 4), "\n")
cat("Observations influentes:", ifelse(length(influential) > 0, 
                                       paste(influential, collapse = ", "), 
                                       "Aucune"), "\n")

if(length(influential) > 0) {
  cat("\nPays influents:\n")
  print(reg_data[influential, c("COUNTRY", "nb_bill", "tax", "log_gdp")])
}


# =============================================================================
# TEST 5 : STABILITÉ DU COEFFICIENT TAX (Leave-One-Out)
# =============================================================================

cat("\n\n=== TEST 5 : STABILITÉ DU COEFFICIENT TAX ===\n")
cat("Teste si le coefficient tax change beaucoup sans chaque pays\n\n")

# Leave-one-out analysis
loo_coefs <- sapply(1:nrow(reg_data), function(i) {
  model_loo <- lm(log_bill ~ tax + log_gdp + fdi + capital, 
                  data = reg_data[-i, ])
  coef(model_loo)["tax"]
})

cat("Coefficient tax original:", round(coef(modele)["tax"], 4), "\n")
cat("Moyenne leave-one-out:", round(mean(loo_coefs), 4), "\n")
cat("Écart-type leave-one-out:", round(sd(loo_coefs), 4), "\n")
cat("Min:", round(min(loo_coefs), 4), "| Max:", round(max(loo_coefs), 4), "\n")
cat("Interprétation:", ifelse(sd(loo_coefs) < 0.01, 
                              "Coefficient stable ✓", 
                              "Coefficient instable ✗"), "\n\n")


# =============================================================================
# TEST 6 : AJOUT DE VARIABLES DE CONTRÔLE (Robustesse)
# =============================================================================

cat("\n=== TEST 6 : ROBUSTESSE AVEC CONTRÔLES SUPPLÉMENTAIRES ===\n\n")

# Ajout de la population
reg_data_pop <- reg_data %>%
  left_join(Population_reduced %>% select(COUNTRY, pop = `2014`), by = "COUNTRY") %>%
  mutate(log_pop = log(pop)) %>%
  filter(!is.na(pop))

model_with_pop <- lm(log_bill ~ tax + log_gdp + fdi + capital + log_pop, 
                     data = reg_data_pop)

cat("Modèle avec population:\n")
print(summary(model_with_pop)$coefficients)

cat("\nComparaison coefficient tax:\n")
cat("Sans population:", round(coef(modele)["tax"], 4), 
    "(p =", round(summary(modele)$coefficients["tax", "Pr(>|t|)"], 4), ")\n")
cat("Avec population:", round(coef(model_with_pop)["tax"], 4), 
    "(p =", round(summary(model_with_pop)$coefficients["tax", "Pr(>|t|)"], 4), ")\n\n")


# =============================================================================
# TEST 7 : EFFET SEUIL DE LA TAXATION
# =============================================================================

cat("\n=== TEST 7 : EXISTE-T-IL UN EFFET SEUIL ? ===\n\n")

# Test avec différents seuils
thresholds <- c(15, 20, 25)

for(threshold in thresholds) {
  reg_data$high_tax <- ifelse(reg_data$tax > threshold, 1, 0)
  model_threshold <- lm(log_bill ~ high_tax + log_gdp + fdi + capital, data = reg_data)
  
  cat("Seuil =", threshold, "%\n")
  cat("  Coefficient high_tax:", round(coef(model_threshold)["high_tax"], 4), "\n")
  cat("  P-value:", round(summary(model_threshold)$coefficients["high_tax", "Pr(>|t|)"], 4), "\n")
  cat("  Significatif:", ifelse(summary(model_threshold)$coefficients["high_tax", "Pr(>|t|)"] < 0.05, 
                                "OUI ✓", "NON ✗"), "\n\n")
}


# =============================================================================
# TEST 8 : EFFET NON-LINÉAIRE (Polynôme)
# =============================================================================

cat("\n=== TEST 8 : EFFET NON-LINÉAIRE DE LA TAXATION ===\n\n")

reg_data$tax_sq <- reg_data$tax^2
model_nonlinear <- lm(log_bill ~ tax + tax_sq + log_gdp + fdi + capital, data = reg_data)

cat("Modèle avec tax²:\n")
print(summary(model_nonlinear)$coefficients[1:3, ])

cat("\nTest de significativité conjointe (tax + tax²):\n")
model_without_tax <- lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)
f_test <- anova(model_without_tax, model_nonlinear)
cat("F-statistic:", round(f_test$F[2], 4), "\n")
cat("P-value:", round(f_test$`Pr(>F)`[2], 4), "\n")
cat("Interprétation:", ifelse(f_test$`Pr(>F)`[2] < 0.05, 
                              "La taxation a un effet significatif (linéaire ou non-linéaire) ✓", 
                              "La taxation n'a pas d'effet significatif ✗"), "\n\n")


# =============================================================================
# TEST 9 : COMPARAISON PAR RÉGIONS
# =============================================================================

cat("\n=== TEST 9 : EFFET DE LA TAXATION PAR RÉGION ===\n\n")

# Ajout des régions (approximation basée sur les noms de pays)
reg_data_region <- reg_data %>%
  mutate(region = case_when(
    COUNTRY %in% c("United States", "Canada", "Mexico") ~ "Amériques",
    COUNTRY %in% c("China", "Japan", "South Korea", "India", "Singapore", 
                   "Hong Kong", "Indonesia", "Malaysia", "Thailand", "Taiwan") ~ "Asie",
    COUNTRY %in% c("Germany", "France", "United Kingdom", "Italy", "Spain", 
                   "Russia", "Switzerland", "Sweden", "Norway", "Netherlands") ~ "Europe",
    TRUE ~ "Autres"
  ))

for(r in unique(reg_data_region$region)) {
  if(sum(reg_data_region$region == r) > 5) {  # Au moins 5 pays
    model_region <- lm(log_bill ~ tax + log_gdp + fdi + capital, 
                       data = reg_data_region %>% filter(region == r))
    cat(r, ":\n")
    cat("  N =", nobs(model_region), "pays\n")
    cat("  Coefficient tax:", round(coef(model_region)["tax"], 4), "\n")
    cat("  P-value:", round(summary(model_region)$coefficients["tax", "Pr(>|t|)"], 4), "\n\n")
  }
}


# =============================================================================
# VISUALISATIONS DIAGNOSTIQUES
# =============================================================================

cat("\n=== VISUALISATIONS DIAGNOSTIQUES ===\n")

# 1. Résidus vs Fitted
plot(modele, which = 1, main = "Résidus vs Valeurs ajustées")

# 2. Q-Q Plot
plot(modele, which = 2, main = "Q-Q Plot (normalité)")

# 3. Scale-Location
plot(modele, which = 3, main = "Scale-Location (homoscédasticité)")

# 4. Cook's distance
plot(modele, which = 4, main = "Cook's distance (influence)")

# 5. Effet partiel de la taxation
plot(reg_data$tax, residuals(lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)),
     xlab = "Taxation (% PIB)", ylab = "Résidus partiels",
     main = "Effet partiel de la taxation (contrôlant pour autres variables)")
abline(lm(residuals(lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)) ~ 
            reg_data$tax), col = "red")


# =============================================================================
# SYNTHÈSE FINALE
# =============================================================================

# =============================================================================
# SYNTHÈSE FINALE
# =============================================================================

cat("\n\n")
cat("═══════════════════════════════════════════════════════════════════\n")
cat("                    SYNTHÈSE DES TESTS                             \n")
cat("═══════════════════════════════════════════════════════════════════\n\n")

cat("QUESTION : La taxation affecte-t-elle la présence de milliardaires ?\n\n")

cat("RÉSULTATS PRINCIPAUX :\n")
cat("• Coefficient taxation:", round(coef(modele)["tax"], 4), "\n")
cat("• P-value:", round(summary(modele)$coefficients["tax", "Pr(>|t|)"], 4), "\n")

# Correction de la ligne problématique
tax_pval <- summary(modele)$coefficients["tax", "Pr(>|t|)"]
if(tax_pval < 0.05) {
  cat("• Significativité: Significatif à 5%\n")
} else if(tax_pval < 0.1) {
  cat("• Significativité: Significatif à 10% (faiblement)\n")
} else {
  cat("• Significativité: Non significatif\n")
}

cat("• R² ajusté:", round(summary(modele)$adj.r.squared, 4), "\n\n")

cat("TESTS DE ROBUSTESSE :\n")
cat("1. Multicolinéarité (VIF) :", ifelse(max(vif_values) < 5, "✓ OK", "✗ Problème"), "\n")
cat("2. Normalité résidus :", ifelse(shapiro_test$p.value > 0.05, "✓ OK", "✗ Problème"), "\n")
cat("3. Homoscédasticité :", ifelse(bp_pvalue > 0.05, "✓ OK", "✗ Problème"), "\n")
cat("4. Observations influentes :", ifelse(length(influential) == 0, "✓ Aucune", 
                                           paste("✗", length(influential), "détectée(s)")), "\n")
cat("5. Stabilité coefficient :", ifelse(sd(loo_coefs) < 0.01, "✓ Stable", "✗ Instable"), "\n\n")

cat("CONCLUSION :\n")
if(tax_pval < 0.05) {
  cat("La taxation a un effet SIGNIFICATIF mais FAIBLE sur la présence de milliardaires.\n")
  cat("L'effet est", ifelse(coef(modele)["tax"] < 0, "NÉGATIF", "POSITIF"), 
      ": +1% de taxation → ", round((exp(coef(modele)["tax"]) - 1) * 100, 2), 
      "% de variation du nombre de milliardaires.\n")
} else if(tax_pval < 0.1) {
  cat("La taxation a un effet FAIBLEMENT SIGNIFICATIF (p < 0.10).\n")
  cat("L'effet existe mais reste MARGINAL par rapport au PIB et aux autres facteurs.\n")
} else {
  cat("La taxation n'a PAS d'effet statistiquement significatif.\n")
  cat("Les OPPORTUNITÉS ÉCONOMIQUES (PIB, investissements) sont les vrais déterminants.\n")
}

cat("\nIMPLICATIONS POUR LA POLITIQUE PUBLIQUE :\n")
cat("→ La taxation seule ne détermine PAS la localisation des milliardaires\n")
cat("→ La TAILLE du marché (GDP) est le facteur dominant\n")
cat("→ Les craintes de 'fuite fiscale massive' sont peut-être exagérées\n")
cat("→ Une politique fiscale équilibrée est possible sans perdre les ultra-riches\n\n")

cat("═══════════════════════════════════════════════════════════════════\n")
```



