---
title: "Final_phase"
format: pdf
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

# Library import

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr) # working with df
library(maps) # use of mpa objects at multiple time
library(viridis) # for better color scale
library(ggrepel)
library(broom)
library(scales)
library(patchwork) # for better graph display
library(ggthemes) # for better graph display
library(lmtest) # heteroscedasticity
library(knitr) # for better df display
library(kableExtra) # for better df display

# Data import

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv", skip = 3)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv", skip = 3)

Capital_formation <- read_csv("Database_project/Capital_formation.csv", skip = 3)

Population <- read_csv("Database_project/Population.csv", skip = 3)


# Cleaning billionaires data set

useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |> # standardization of names
  rename(COUNTRY = citizenship)

Billionaires_1996 <-
  Billionaires |>
  filter(year == "1996")

Billionaires_2001 <-
  Billionaires |>
  filter(year == "2001")

Billionaires_2014 <-
  Billionaires |>
  filter(year == "2014")

Billionaires_wide <-
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Cleaning the IMF dataset

Country <-
  Billionaires |>
  distinct(COUNTRY)

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- list(
  "Bahrain, Kingdom of" ~ "Bahrain",
  "China, People's Republic of" ~ "China",
  "Egypt, Arab Republic of" ~ "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" ~ "Hong Kong",
  "Kazakhstan, Republic of" ~ "Kazakhstan",
  "Liechtenstein, Principality of" ~ "Liechtenstein",
  "Lithuania, Republic of" ~ "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" ~ "Macau",
  "Netherlands, The" ~ "Netherlands",
  "Poland, Republic of" ~ "Poland",
  "Russian Federation" ~ "Russia",
  "Korea, Republic of" ~ "South Korea",
  "Eswatini, Kingdom of" ~ "Swaziland",
  "Taiwan Province of China" ~ "Taiwan",
  "Tanzania, United Republic of" ~ "Tanzania",
  "Türkiye, Republic of" ~ "Turkey",
  "Venezuela, República Bolivariana de" ~ "Venezuela"
)



IMF <-
  IMF |>
  mutate(COUNTRY = case_match(COUNTRY, !!!map_IMF, .default = COUNTRY))

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Cleaning the WB datasets

WB_names <-
  FDI |>
  distinct(`Country Name`)

map_WB <- list(
  "Czech Republic" ~ "Czechia",
  "Egypt, Arab Rep." ~ "Egypt",
  "Hong Kong SAR, China" ~ "Hong Kong",
  "Macao SAR, China" ~ "Macau",
  "Russian Federation" ~ "Russia",
  "Korea, Rep." ~ "South Korea",
  "Venezuela, RB" ~ "Venezuela",
  "Viet Nam" ~ "Vietnam"
)

FDI <-
  FDI |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Population <-
  Population |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)
```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) State of the world of billionaires in 2014

### 1.1 Global mapping of billionaires

```{r}
#| label: Mapping graoh of billionaires
#| echo: false
#| message: false
#| warning: false

# Aggregation by country: number and total wealth

billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data

world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format

country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
    country_mapping[COUNTRY], # Apply mapping if found
    COUNTRY
  )) # Keep original name otherwise

# Join billionaire data with map data

world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country

map_count <- ggplot() +
  geom_polygon(
    data = world_map, # Draw all countries as base layer
    aes(x = long, y = lat, group = group), # Define coordinates and grouping
    fill = "grey90", color = "white", linewidth = 0.1
  ) + # Draw only countries with billionaires
  geom_polygon(
    data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
    aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
    color = "white", linewidth = 0.1
  ) +
  scale_fill_viridis(
    option = "plasma",
    name = "Number of\nbillionaires",
    trans = "log10", # Apply logarithmic transformation
    breaks = c(1, 5, 10, 50, 100, 500),
    direction = -1
  ) +
  theme_void() + # Remove axes and background for clean map
  labs(title = "Global distribution of billionaires in 2014") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```

For this specific map, we used AI assistance to help refine the R code and generate a clearer, more visual representation. We already had the initial intuition and the base approach, but we chose this mapping option because it is easier to read and more informative than a standard “classic” ggplot output.

This map shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs.


### 1.2 Caracteristics and origin of billionaires' wealth 

```{r}
#| label: Distribution of billionaires graph
#| echo: false
#| message: false
#| warning: false

Sectors <-
  Billionaires_2014 |>
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |>
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(
    cols = c(nb_self_made, nb_inherited),
    names_to = "origine",
    values_to = "count"
  ) |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


Plot_sector <- ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Distribution of billionaires by sector and origin of wealth in 2014",
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  ) +
  theme_clean(base_size = 10)

Source <-
  Billionaires |>
  drop_na(selfmade) |>
  summarise(n = n(), .by = c(selfmade, year))

Plot_year <- ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(
    title = "Source of wealth by year",
    y = "Year",
    x = "Number",
    fill = "Origin"
  ) +
  theme_clean(base_size = 10)

Plot_sector / Plot_year
```

The first graph illustrates the distribution of billionaires across various industries, highlighting that real estate is the most prominent sector for amassing such wealth, followed by retail and construction. A significant trend across almost all sectors is the clear dominance of self-made individuals over those who inherited their fortunes. This disparity is particularly pronounced in modern industries like technology, software, and hedge funds, where inherited wealth is virtually non-existent, whereas more traditional sectors like oil and banking show a relatively higher presence of generational wealth.

The second graph tracks the evolution of billionaire wealth origins from 1996 to 2014, revealing a dramatic surge in the total number of billionaires globally, which increased from approximately 400 to over 1,600. While the billionaire population was much smaller and more evenly divided between inherited and self-made origins in 1996, the landscape shifted significantly by 2014. The expansion of this demographic over the eighteen-year period has been overwhelmingly driven by self-made entrepreneurs, who now constitute the vast majority of the total, indicating a global shift toward new wealth creation rather than the simple maintenance of inherited estates.

## 2) First model

```{r}
#| label: First model
#| message: false
#| warning: false
#| echo: false

# Data preparation

reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |>
    select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |>
    select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |>
    select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |>
    select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

# First model

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)
summary(modele)
```

Our baseline regression links the (log) number of billionaires in a country to our taxation proxy (tax revenue), economic size (GDP), openness (FDI), and investment capacity (capital formation). Overall, the model fits the data well (Adjusted R²= 0.771; F-test p-value < 10^(-14)), suggesting these macroeconomic factors explain a large share of cross-country variation in billionaire presence.

The coefficient on tax revenue is positive (β = 0.022) and marginally significant at the 10% level (p = 0.097). Interpreting tax revenue as a proxy for the overall level of taxation/fiscal pressure, this result suggests that countries with stronger tax capacity tend to host more, not fewer, billionaires. Economically, this is consistent with the idea that higher-tax countries often also provide the institutional and public-good environment that supports wealth creation and wealth retention—stable legal systems, infrastructure, education, and high-quality public services.

Therefore, taxation does not automatically translate into a weaker ability to attract or maintain top wealth holders. In the context of an open economy, this weak but positive association also suggests that “competitiveness” is not solely driven by low taxation: some high-tax jurisdictions can remain attractive because they compensate with stability and market opportunities.

The strongest determinant is economic scale: log GDP is highly significant and large. This implies that larger economies systematically host more billionaires, which aligns with the idea that big markets generate more high-value firms, deeper capital markets, and more opportunities for large fortunes. This is important for our research question because it highlights that billionaire wealth concentration is strongly linked to market size and overall economic performance—taxation effects must be interpreted conditional on these structural fundamentals.

FDI is also positive and significant, indicating that more open and internationally connected economies tend to have more billionaires. This supports the “national competitiveness in an open economy” channel: countries that attract foreign investment and integrate into global capital flows are also those where billionaire wealth is more prevalent. In other words, openness and investment attractiveness appear complementary to the presence of ultra-wealthy individuals rather than substitutable.

Finally, capital formation is not statistically significant (p = 0.40), suggesting that once GDP size and FDI are accounted for, differences in domestic investment intensity (as measured here) do not add much explanatory power for billionaire counts.

Taken together, this baseline model suggests that, using tax revenue as a proxy for taxation, higher taxation is not associated with a lower billionaire presence, while economic performance (GDP) and openness/competitiveness (FDI) are strongly associated with billionaire concentration. This supports an interpretation where taxation can coexist with competitiveness in open economies, provided the broader institutional and market conditions remain favorable.

### 2.1 Multicolinearity analysis

We here test for multicollinearity among the explanatory variables using the Variance Inflation Factor (VIF).

```{r}
#| label: Multicolinearity analysis
#| message: false
#| warning: false
#| echo: false

library(car) # vif, setting it here to avoid conflict with dplyr

vif_df <- data.frame(
  Variable = names(vif(modele)),
  VIF = as.numeric(vif(modele))
)
vif_df
```

The VIF values (≈1.19–1.34) are close to 1, indicating no meaningful multicollinearity among the regressors, so coefficient significance is unlikely to be driven by correlated explanatory variables.


### 2.2 Normality test for residuals

Here, we test whether our data are distributed according to a normal distribution. To do this, we calculate the Shapiro-Wilk normality test statistics. We obtain the following results :

```{r}
#| label: Normality test for residuals
#| message: false
#| warning: false
#| echo: false

shapiro_test <- shapiro.test(residuals(modele))

shapiro_test
```

- The p-value is 0.7615 :  It is well above the critical threshold of 0.05 (or 5%), which means that we do not reject the null hypothesis (H0). We conclude that the residuals of our model follow a normal distribution.
- The points (our data) follow the reference line almost perfectly, there are no major deviations, either at the extremes (no serious outliers) or in the centre. The distribution is symmetrical.

### 2.3 Heteroscedasticity 

We here test for heteroscedasticity using the Breusch–Pagan test to check whether the error variance is constant. 

```{r}
#| label: Heteroscedasticy
#| message: false
#| warning: false
#| echo: false

bp <- bptest(modele)


bp_df <- data.frame(
  Test = "Studentized Breusch–Pagan",
  Test_statistic = as.numeric(bp$statistic),
  p_value = as.numeric(bp$p.value)
)

bp_df
```

With a p-value of 0.7439, we fail to reject the null hypothesis of constant error variance, meaning there is no evidence of heteroskedasticity in the model.

### 2.4 Conclusion
Overall, the diagnostic tests support the validity of the OLS assumptions for our first model.
The VIF results (around 1.2–1.3) indicate no multicollinearity, so the regressors are not strongly correlated.
The Shapiro–Wilk test (p = 0.7615) suggests the residuals are approximately normally distributed.
The Breusch–Pagan test (p = 0.7439) shows no evidence of heteroskedasticity, meaning the error variance can be treated as constant.
Taken together, these checks imply that our coefficient estimates and standard inference from this model are generally reliable.

## 3) Modelling tax evasion by billionaires
### 3.1 Creation of a flight-risk score

After completing our first model, we decided to develop a flight risk score. The aim of our approach is to establish, for each billionaire the risk of flight from their current national territory if a tax is introduced. With this in mind, our tax evasion risk score is based on four distinct components :

- A tax burden score between 0 and 1 calculated according to the following standardization : $\frac{(\text{tax in country i})- (\text{min tax})}{(\text{max tax})- (\text{min tax})}$. This score is thus normalized between 0 (minimum taxation) and 1 (maximum taxation). We consider that higher taxation would encourage more tax evasion.  
- Whether the billionaire's main activity belongs to an industry considered to be internationally mobile. Among the industries in the original dataset, we have identified the following as mobile : "Diversified financial",  "Technology-Computer", "Hedge funds", "Private equity/leverage buyout",      "Venture capital", "Media", "Money Management". Belonging to one of this industry gives a score of 1 and 0 otherwise as we believe that working in such a sector would facilitate mobility for the billionaire. 
- The source of the billionaire's wealth, with a score of 1 if "self-made" and 0 if "inherited". This score is used to capture the idea that inherited wealth is more likely to be accompanied by a strong attachment to the territory of origin. The idea is that inherited wealth tends to involve real estate and historically established industries, which provide incentives to stay put. 
- Proximity to a tax haven. Based on the official definition, we first drew up the following list of tax havens: "Switzerland", "Singapore", "Hong Kong", "Liechtenstein", "Monaco", "Luxembourg", "United Arab Emirates" and "Bermuda". We then identified the countries considered close enough to a tax haven to provide billionaires with a significant incentive to relocate. To do this, we took into account membership of the Schengen area for European countries (where mobility to the border of the tax haven is facilitated, whether for Switzerland, Luxembourg or Liechtenstein) and the sharing of a coastline with the same ocean for Asian and Caribbean tax havens. This gives us a score of 1 if close, 0 otherwise.

```{r}
#| label: Creation of the 4 scores
#| echo: false
#| message: false
#| warning: false


# Sectors considered "mobile"

mobile_sectors <- c("Diversified financial", "Technology-Computer", "Hedge funds", "Private equity/leverage buyout", "Venture capital", "Media", "Money Management")

# List of tax havens/refuges

tax_havens <- c(
  "Switzerland", "Singapore", "Hong Kong", "Liechtenstein",
  "Monaco", "Luxembourg", "United Arab Emirates", "Bermuda"
)

# Calcul of the 4 indicators

flight_risk_index <-
  Billionaires_2014 |>
  left_join(Tax_revenue_reduced |> select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  mutate(
    # tax burden score calculation
    tax_normalized =
      ifelse(tax != 0, (tax - min(tax, na.rm = TRUE)) / (max(tax, na.rm = TRUE) - min(tax, na.rm = TRUE)), 0),
    tax_normalized =
      ifelse(is.na(tax) == TRUE, 0, tax_normalized),
    # mobility score calculation
    mobile_sector = ifelse(industry %in% mobile_sectors, 1, 0),
    # source of wealth index computation
    selfmade = ifelse(is.na(selfmade) == TRUE, "self-made", selfmade),
    is_selfmade = ifelse(selfmade == "self-made", 1, 0),
    is_selfmade = replace_na(is_selfmade, 0),

    # proximity to tax haven index computation
    near_haven = case_when(
      COUNTRY %in% c(
        "France", "Germany", "Italy", "Spain", "United Kingdom",
        "Belgium", "Netherlands", "Austria", "Sweden", "Norway",
        "Portugal", "Denmark", "Czech Republic", "Finland", "Greece",
        "Poland", "Romania", "Cyprus", "Lithuania"
      ) ~ 1,
      COUNTRY %in% c(
        "China", "Japan", "South Korea", "India", "Malaysia",
        "Thailand", "Indonesia", "Taiwan", "Philippines", "Macau", "Vietnam"
      ) ~ 1,
      COUNTRY %in% c(
        "Saudi Arabia", "Kuwait", "Israel", "Turkey", "Egypt",
        "Lebanon", "Oman"
      ) ~ 1,
      COUNTRY %in% c(
        "United States", "Canada", "Brazil", "Mexico", "Colombia",
        "Venezuela", "St. Kitts and Nevis"
      ) ~ 1,
      TRUE ~ 0
    ),
    in_tax_haven = ifelse(COUNTRY %in% tax_havens, 1, 0)
  )
```

For each billionaire, four indicators are calculated, which are then combined to obtain the final flight risk score. For the calculation, we assigned the following coefficients :

$$
\text{Score} = 0.45 \cdot \text{Taxation} + 0.3 \cdot \text{Mobility} + 0.15 \cdot \text{Source of wealth} + 0.1 \cdot \text{Proximity}
$$

- Taxation has the highest coefficient because it is theoretically the most significant incentive to leave.
- Mobility comes next, as it represents the most difficult physical constraint to overcome.
- The origin of wealth has a less significant coefficient because its impact remains limited due to the complexity of accurately establishing a billionaire's personal ties to their territory.
- The smallest coefficient is associated with proximity to a tax haven. Despite the physical constraint that distance represents, the economy remains globalized and billionaires have ample means to circumvent the issue.

Tax havens are assigned a default score of 0, and for the sake of clarity, categories are assigned according to the score: "Low" if below 0.3, “Medium” between 0.3 and 0.6, and "High" if above 0.6.

```{r}
#| label: Aggregation of the 4 scores
#| echo: false
#| message: false
#| warning: false

flight_risk_index <-
  flight_risk_index |>
  mutate(
    wealth_bil = networthusbillion / 10, # Conversion of tenths of billions into billions
    flight_risk_score = (tax_normalized * 0.45 +
      mobile_sector * 0.30 +
      is_selfmade * 0.15 +
      near_haven * 0.10),
    flight_risk_score = ifelse(in_tax_haven == 1, 0, flight_risk_score),
    risk_category = case_when(
      flight_risk_score < 0.3 ~ "Low",
      flight_risk_score >= 0.3 & flight_risk_score < 0.6 ~ "Medium",
      flight_risk_score >= 0.6 ~ "High"
    )
  )
```

### 3.2 Conversion of the flight-score into a probability using a logistic transformation

We convert the flight-risk score into an expected flight probability using a logistic transformation because the score is an index (0–1) that ranks individuals by relative exposure to mobility incentives, but it is not directly interpretable as a probability or a share of wealth that would relocate. The logistic mapping ensures predicted values remain bounded between 0 and 1 and allows for a more realistic non-linear response: small changes in risk matter little for low-risk profiles, while effects become stronger for high-risk profiles without mechanically producing extreme flight rates.

Since we do not observe actual relocation decisions in our data, we calibrate the logistic parameters $(\alpha)$ and $(\beta)$ transparently using two anchor points from the score distribution. 
Let $(p_i=\text{plogis}(\alpha+\beta,\text{score}*i))$. We set the 10th percentile of the score $((s*{10}))$ to correspond to a low expected flight rate $(p_{10}=2\%)$, and the 90th percentile $((s_{90}))$ to correspond to a high expected flight rate $(p_{90}=25\%)$. These values reflect a “moderate mobility” scenario: low-risk individuals have only a small chance to relocate, while high-risk individuals display a substantially higher—but still bounded—propensity to move. The resulting $(\alpha)$ and $(\beta)$ are then uniquely determined by matching $(p(s_{10})=0.02)$ and $(p(s_{90})=0.25)$, and this calibration yields an average implied flight probability of about 10.8% in the sample.

We could test the model according to the type of calibration we want :

- "Moderate" calibration: (2% → 25%) used here. We calibrate the logistic transformation by requiring that individuals in the 10th percentile of the score have an expected leakage of approximately 2%, and those in the 90th percentile have an expected leakage of 25%. This gives an average probability of approximately 10.8%
- "Low" calibration: (1% → 15%)
- "High" calibration: (3% → 35%)


We used AI assistance to refine and validate the implementation of the logit-based calibration code. 



```{r}
#| label: Conversion using the logistic function
#| echo: false
#| message: false
#| warning: false


logit <- function(p) log(p / (1 - p)) # Define the logit function: logit(p) = log(p / (1 - p))
# two anchor points from the empirical distribution of the score:
s_low <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.10, na.rm = TRUE)) #    s_low  = 10th percentile of flight_risk_score (low-risk profiles)
s_high <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.90, na.rm = TRUE)) #    s_high = 90th percentile of flight_risk_score (high-risk profiles)

# Target flight probabilities at these anchors (moderate mobility scenario):
# p_low  is the expected flight probability at s_low
# p_high is the expected flight probability at s_high
p_low <- 0.02 # ~2% expected flight at the low-risk (10th percentile) score
p_high <- 0.25 # ~25% expected flight at the high-risk (90th percentile) score

# Solve for beta and alpha in:
#    p(score) = plogis(alpha + beta * score)
# Using the logit transform:
#    logit(p_low)  = alpha + beta * s_low
#    logit(p_high) = alpha + beta * s_high
# Subtracting the two equations gives beta, then alpha follows.
beta <- (logit(p_high) - logit(p_low)) / (s_high - s_low)
alpha <- logit(p_low) - beta * s_low

# Compute the implied flight probability for each billionaire:
flight_risk_index <- 
  flight_risk_index |>
  mutate(
    flight_prob = ifelse(flight_risk_score!=0, plogis(alpha + beta * flight_risk_score), 0)
  )
```

### 3.3 Analysis of the results

Now that we have converted the flight-risk score into a calibrated flight probability (flight_prob), we aggregate the results at the country level. This allows us to compare countries in terms of (i) billionaire concentration, (ii) total billionaire wealth, and (iii) both the average risk index and the average implied mobility (probability), focusing on countries with a sufficiently large billionaire sample.

```{r}
#| label: Preparing the results' display
#| echo: false
#| message: false
#| warning: false



country_risk <-
  flight_risk_index |>
  summarise(
    nb_billionaires = n(), # number of billionaires
    total_wealth = sum(wealth_bil, na.rm = TRUE), # total wealth ($bn), unit-corrected
    avg_flight_risk = mean(flight_risk_score, na.rm = TRUE), # average risk score (index)
    avg_flight_prob = mean(flight_prob, na.rm = TRUE), # average risk score (prob)
    nb_high_risk = sum(risk_category == "High", na.rm = TRUE), # average calibrated flight probability
    pct_high_risk = (nb_high_risk / n()) * 100, # % of high-risk individuals
    tax = mean(tax, na.rm = TRUE), # average tax proxy (same within a country)
    .by = COUNTRY
  ) |>
  arrange(desc(avg_flight_risk))

top_country_risk <-
  country_risk |>
  filter(nb_billionaires > 10) |>
  slice_max(order_by = avg_flight_prob, n = 10)

top_country_risk |>
  select(1, 2, 5) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    longtable = TRUE,
    caption = "Top Country Risk Analysis",
    digits = 2
  )

```

The table aggregates the flight-risk indicators at the country level for countries with more than ten billionaires. It reports the number of billionaires, total billionaire wealth (in $bn), the average flight-risk score (our composite index), the average implied flight probability (from the calibrated logistic mapping), and the share of individuals classified as “high risk”.

Overall, the ranking is economically coherent: countries with higher average flight-risk scores also display higher average flight probabilities, confirming that the logistic calibration preserves the ordering of the underlying index while translating it into an interpretable mobility metric. At the top of the distribution, countries such as the United Kingdom, Israel, and Sweden exhibit the highest average risk scores and correspondingly high predicted mobility. This is consistent with our mechanisms: a greater presence of internationally mobile sectors and lower frictions to cross-border relocation (captured by our proximity proxy) increase the expected responsiveness of billionaire wealth to a new wealth tax.

France appears in the upper-middle of the distribution, with a moderate-to-high average risk score and an implied mobility rate around the low-teens. This suggests that, while fiscal pressure can create incentives to relocate, the average predicted response remains limited by the composition and embeddedness of wealth (e.g., fortunes tied to less footloose activities or inherited structures). In contrast, the United States concentrates by far the largest number of billionaires and the largest total wealth, yet its average risk score is not among the highest. This result supports the idea that market size and domestic opportunities can reduce the need to relocate even in an open economy: large economies can host many billionaires without necessarily exhibiting the highest average mobility.

Finally, the “% high-risk” column highlights within-country heterogeneity. Some countries have a relatively high share of high-risk profiles, indicating that potential tax-base erosion may be concentrated in a subset of individuals rather than uniformly spread. This distinction is important for policy simulations: revenue losses depend not only on average mobility but also on how risk is distributed across billionaire profiles and, especially, across the largest fortunes.




## 4) Scenario testing : the impact of new taxation 

```{r}
#| label: Define G7 and G20 members
#| message: false
#| warning: false
#| echo: false

G7 <- c(
  "United States", "United Kingdom", "Germany", "France",
  "Italy", "Canada", "Japan"
)

G20 <- c(
  G7, "China", "India", "Brazil", "Russia", "Australia",
  "South Korea", "Mexico", "Indonesia", "Turkey", "Saudi Arabia",
  "Argentina", "South Africa"
)
```


```{r}
#| message: false
#| warning: false
#| echo: false
## CHUNK 2.2 : Fonction de simulation des recettes fiscales
# -----------------------------------------------------------------------------
# Cette fonction calcule les recettes d'une taxe en tenant compte de la fuite
# Elle prend en entrée un coefficient d'ajustement de la probabilité de fuite

simulate_tax_scenario <- function(data, scenario_name, flight_multiplier) {
  # flight_multiplier : ajuste la probabilité de fuite
  # = 1.0 pour fuite normale, < 1.0 pour fuite réduite (coordination)

  results <- data |>
    mutate(
      effective_flight_prob = flight_prob * flight_multiplier, # Probabilité effective de fuite dans ce scénario
      effective_flight_prob = pmin(effective_flight_prob, 1), # Max = 1
      initial_wealth = wealth_bil,
      fled_wealth = initial_wealth * effective_flight_prob, # Richesse qui fuit à l'étranger
      taxable_wealth = initial_wealth - fled_wealth, # Richesse restante (qui sera taxée)
      potential_revenue = initial_wealth * 0.02, # Recettes potentielles (si aucune fuite)
      actual_revenue = taxable_wealth * 0.02, # Recettes réelles (après fuite)
      lost_revenue = potential_revenue - actual_revenue, # Perte de recettes due à la fuite
    )


  # Résumé global
  global_summary <- results |>
    summarise(
      scenario = scenario_name,
      total_billionaires = n(),
      avg_flight_prob = mean(effective_flight_prob) * 100,
      total_wealth_billion = sum(initial_wealth),
      fled_wealth_billion = sum(fled_wealth),
      potential_revenue_billion = sum(potential_revenue),
      actual_revenue_billion = sum(actual_revenue),
      lost_revenue_billion = sum(lost_revenue),
      collection_efficiency = (actual_revenue_billion / potential_revenue_billion) * 100
    )

  # Résumé par pays (top contributeurs)
  country_summary <- results |>
    group_by(COUNTRY) |>
    summarise(
      scenario = scenario_name,
      nb_bill = n(),
      potential = sum(potential_revenue),
      actual = sum(actual_revenue),
      lost = sum(lost_revenue)
    ) |>
    arrange(desc(actual))

  return(list(global = global_summary, by_country = country_summary))
}
```



```{r}

#| label: Test de scénarios
#| echo: false
#| message: false
#| warning: false

sc_fr <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY == "France"), "France seule", 1.5)
sc_g7 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G7), "Coordination G7", 0.9)
sc_g20 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G20), "Coordination G20", 0.7)

summary_table_all <- bind_rows(sc_fr$global, sc_g7$global, sc_g20$global) |>
  transmute(
    Scenario = scenario,
    `# billionaires` = total_billionaires,
    `Expected flight rate (%)` = round(avg_flight_prob, 1),
    `Potential revenue ($bn)` = round(potential_revenue_billion, 3),
    `Actual revenue ($bn)` = round(actual_revenue_billion, 3),
    `Lost revenue ($bn)` = round(lost_revenue_billion, 3),
    `Collection efficiency (%)` = round(collection_efficiency, 1)
  )

summary_table_all
```

concave loss of revenue 

```{r}
#| label: Top 5 contributeurs à la perte
#| message: false
#| warning: false
#| echo: false
top_loss_fr <- flight_risk_index |>
  filter(COUNTRY == "France") |>
  mutate(
    p_france = pmin(flight_prob * 1.5, 1),
    potential_rev = wealth_bil * 0.02,
    actual_rev = wealth_bil * (1 - p_france) * 0.02,
    lost_rev = potential_rev - actual_rev
  ) |>
  arrange(desc(lost_rev)) |>
  select(name, wealth_bil, p_france, lost_rev, industry, selfmade) |>
  head(10)

top_loss_fr
```

```{r}
#| message: false
#| warning: false
#| echo: false

fr_detail <- flight_risk_index |>
  filter(COUNTRY == "France") |>
  mutate(
    eff = pmin(flight_prob * 1.5, 1),
    lost = wealth_bil * 0.02 * eff
  ) |>
  arrange(desc(wealth_bil)) |>
  mutate(
    cum_lost_share = cumsum(lost) / sum(lost),
    rank = row_number(),
    pct = rank / n()
  )

# Part des pertes expliquée par le top 10% (les plus riches)
fr_detail |>
  summarise(share_lost_top10 = max(cum_lost_share[pct <= 0.10]))
```
