---
title: "Final_phase"
format: pdf
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

# Library import

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr) # working with df
library(maps) # use of mpa objects at multiple time
library(viridis) # for better color scale
library(ggrepel)
library(broom)
library(scales)
library(patchwork) # for better graph display
library(ggthemes) # for better graph display
library(lmtest) # heteroscedasticity
library(knitr) # for better df display
library(kableExtra) # for better df display
library(modelsummary)

# Data import

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
  delim = ";", escape_double = FALSE, trim_ws = TRUE
)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv", skip = 3)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv", skip = 3)

Capital_formation <- read_csv("Database_project/Capital_formation.csv", skip = 3)

Population <- read_csv("Database_project/Population.csv", skip = 3)


# Cleaning billionaires data set

useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |> # standardization of names
  rename(COUNTRY = citizenship)

Billionaires_1996 <-
  Billionaires |>
  filter(year == "1996")

Billionaires_2001 <-
  Billionaires |>
  filter(year == "2001")

Billionaires_2014 <-
  Billionaires |>
  filter(year == "2014")

Billionaires_wide <-
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Cleaning the IMF dataset

Country <-
  Billionaires |>
  distinct(COUNTRY)

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- list(
  "Bahrain, Kingdom of" ~ "Bahrain",
  "China, People's Republic of" ~ "China",
  "Egypt, Arab Republic of" ~ "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" ~ "Hong Kong",
  "Kazakhstan, Republic of" ~ "Kazakhstan",
  "Liechtenstein, Principality of" ~ "Liechtenstein",
  "Lithuania, Republic of" ~ "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" ~ "Macau",
  "Netherlands, The" ~ "Netherlands",
  "Poland, Republic of" ~ "Poland",
  "Russian Federation" ~ "Russia",
  "Korea, Republic of" ~ "South Korea",
  "Eswatini, Kingdom of" ~ "Swaziland",
  "Taiwan Province of China" ~ "Taiwan",
  "Tanzania, United Republic of" ~ "Tanzania",
  "Türkiye, Republic of" ~ "Turkey",
  "Venezuela, República Bolivariana de" ~ "Venezuela"
)



IMF <-
  IMF |>
  mutate(COUNTRY = case_match(COUNTRY, !!!map_IMF, .default = COUNTRY))

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Cleaning the WB datasets

WB_names <-
  FDI |>
  distinct(`Country Name`)

map_WB <- list(
  "Czech Republic" ~ "Czechia",
  "Egypt, Arab Rep." ~ "Egypt",
  "Hong Kong SAR, China" ~ "Hong Kong",
  "Macao SAR, China" ~ "Macau",
  "Russian Federation" ~ "Russia",
  "Korea, Rep." ~ "South Korea",
  "Venezuela, RB" ~ "Venezuela",
  "Viet Nam" ~ "Vietnam"
)

FDI <-
  FDI |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Population <-
  Population |>
  mutate(`Country Name` = case_match(`Country Name`, !!!map_WB, .default = `Country Name`))

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)
```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) State of the world of billionaires in 2014

### 1.1 Global mapping of billionaires

```{r}
#| label: Mapping graoh of billionaires
#| echo: false
#| message: false
#| warning: false

# Aggregation by country: number and total wealth

billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data

world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format

country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
    country_mapping[COUNTRY], # Apply mapping if found
    COUNTRY
  )) # Keep original name otherwise

# Join billionaire data with map data

world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country

map_count <- ggplot() +
  geom_polygon(
    data = world_map, # Draw all countries as base layer
    aes(x = long, y = lat, group = group), # Define coordinates and grouping
    fill = "grey90", color = "white", linewidth = 0.1
  ) + # Draw only countries with billionaires
  geom_polygon(
    data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
    aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
    color = "white", linewidth = 0.1
  ) +
  scale_fill_viridis(
    option = "plasma",
    name = "Number of\nbillionaires",
    trans = "log10", # Apply logarithmic transformation
    breaks = c(1, 5, 10, 50, 100, 500),
    direction = -1
  ) +
  theme_void() + # Remove axes and background for clean map
  labs(title = "Global distribution of billionaires in 2014") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```

For this specific map, we used AI assistance to help refine the R code and generate a clearer, more visual representation. We already had the initial intuition and the base approach, but we chose this mapping option because it is easier to read and more informative than a standard “classic” ggplot output.

This map shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs.


### 1.2 Caracteristics and origin of billionaires' wealth 

```{r}
#| label: fig-Distribution-of-billionaires
#| fig-cap: "Distribution of billionaires by sector and origin of wealth in 2014"
#| echo: false
#| message: false
#| warning: false

Sectors <-
  Billionaires_2014 |>
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |>
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(
    cols = c(nb_self_made, nb_inherited),
    names_to = "origine",
    values_to = "count"
  ) |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


Plot_sector <- ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  ) +
  theme_clean(base_size = 10)

Source <-
  Billionaires |>
  drop_na(selfmade) |>
  summarise(n = n(), .by = c(selfmade, year))

Plot_year <- ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(
    title = "Source of wealth by year",
    y = "Year",
    x = "Number",
    fill = "Origin"
  ) +
  theme_clean(base_size = 10)

Plot_sector / Plot_year
```

The first graph illustrates the distribution of billionaires across various industries, highlighting that real estate is the most prominent sector for amassing such wealth, followed by retail and construction. A significant trend across almost all sectors is the clear dominance of self-made individuals over those who inherited their fortunes. This disparity is particularly pronounced in modern industries like technology, software, and hedge funds, where inherited wealth is virtually non-existent, whereas more traditional sectors like oil and banking show a relatively higher presence of generational wealth.

The second graph tracks the evolution of billionaire wealth origins from 1996 to 2014, revealing a dramatic surge in the total number of billionaires globally, which increased from approximately 400 to over 1,600. While the billionaire population was much smaller and more evenly divided between inherited and self-made origins in 1996, the landscape shifted significantly by 2014. The expansion of this demographic over the eighteen-year period has been overwhelmingly driven by self-made entrepreneurs, who now constitute the vast majority of the total, indicating a global shift toward new wealth creation rather than the simple maintenance of inherited estates.

## 2) First model

```{r}
#| label: First model
#| message: false
#| warning: false
#| echo: false

# Data preparation

reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |>
    select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |>
    select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |>
    select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |>
    select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

# First model

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)
modelsummary(
  modele,
  estimate  = "{estimate}{stars} ({std.error})", 
  statistic = NULL, 
  gof_omit = "AIC|BIC|Log.Lik|F|RMSE",
  output = "kableExtra" 
)
```

Our baseline regression links the (log) number of billionaires in a country to our taxation proxy (tax revenue), economic size (GDP), openness (FDI), and investment capacity (capital formation). Overall, the model fits the data well (Adjusted R²= 0.771; F-test p-value < 10^(-14)), suggesting these macroeconomic factors explain a large share of cross-country variation in billionaire presence.

The coefficient on tax revenue is positive (β = 0.022) and marginally significant at the 10% level (p = 0.097). Interpreting tax revenue as a proxy for the overall level of taxation/fiscal pressure, this result suggests that countries with stronger tax capacity tend to host more, not fewer, billionaires. Economically, this is consistent with the idea that higher-tax countries often also provide the institutional and public-good environment that supports wealth creation and wealth retention—stable legal systems, infrastructure, education, and high-quality public services.

Therefore, taxation does not automatically translate into a weaker ability to attract or maintain top wealth holders. In the context of an open economy, this weak but positive association also suggests that “competitiveness” is not solely driven by low taxation: some high-tax jurisdictions can remain attractive because they compensate with stability and market opportunities.

The strongest determinant is economic scale: log GDP is highly significant and large. This implies that larger economies systematically host more billionaires, which aligns with the idea that big markets generate more high-value firms, deeper capital markets, and more opportunities for large fortunes. This is important for our research question because it highlights that billionaire wealth concentration is strongly linked to market size and overall economic performance—taxation effects must be interpreted conditional on these structural fundamentals.

FDI is also positive and significant, indicating that more open and internationally connected economies tend to have more billionaires. This supports the “national competitiveness in an open economy” channel: countries that attract foreign investment and integrate into global capital flows are also those where billionaire wealth is more prevalent. In other words, openness and investment attractiveness appear complementary to the presence of ultra-wealthy individuals rather than substitutable.

Finally, capital formation is not statistically significant (p = 0.40), suggesting that once GDP size and FDI are accounted for, differences in domestic investment intensity (as measured here) do not add much explanatory power for billionaire counts.

Taken together, this baseline model suggests that, using tax revenue as a proxy for taxation, higher taxation is not associated with a lower billionaire presence, while economic performance (GDP) and openness/competitiveness (FDI) are strongly associated with billionaire concentration. This supports an interpretation where taxation can coexist with competitiveness in open economies, provided the broader institutional and market conditions remain favorable.

### 2.1 Multicolinearity analysis

We here test for multicollinearity among the explanatory variables using the Variance Inflation Factor (VIF).

```{r}
#| label: tbl-vif
#| tbl-cap: "Multicolinearity analysis (VIF)"
#| message: false
#| warning: false
#| echo: false

library(car)

vif_df <- data.frame(
  Variable = names(vif(modele)),
  VIF = as.numeric(vif(modele))
)

kable(vif_df,
      col.names = c("Independent Variable", "VIF"),
      digits = 3,
      booktabs = TRUE) |> 
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE)
```

The VIF values (≈1.19–1.34) are close to 1, indicating no meaningful multicollinearity among the regressors, so coefficient significance is unlikely to be driven by correlated explanatory variables.


### 2.2 Normality test for residuals

Here, we test whether our data are distributed according to a normal distribution. To do this, we calculate the Shapiro-Wilk normality test statistics. We obtain the following results :

```{r}
#| label: tbl-shapiro
#| tbl-cap: "Shapiro-Wilk"
#| message: false
#| warning: false
#| echo: false

shapiro_test <- tidy(shapiro.test(residuals(modele)))

kable(shapiro_test,
      col.names = c("W", "p-value", "Méthode"),
      digits = 4,
      booktabs = TRUE) |> 
  kable_styling(latex_options = "hold_position")
```

- The p-value is 0.7615 :  It is well above the critical threshold of 0.05 (or 5%), which means that we do not reject the null hypothesis (H0). We conclude that the residuals of our model follow a normal distribution.
- The points (our data) follow the reference line almost perfectly, there are no major deviations, either at the extremes (no serious outliers) or in the centre. The distribution is symmetrical.

### 2.3 Heteroscedasticity 

We here test for heteroscedasticity using the Breusch–Pagan test to check whether the error variance is constant. 

```{r}
#| label: tbl-breusch-pagan
#| tbl-cap: "Breusch-Pagan"
#| message: false
#| warning: false
#| echo: false

bp <- bptest(modele)

bp_df <- data.frame(
  Test = "Studentized Breusch-Pagan",
  Test_statistic = as.numeric(bp$statistic),
  p_value = as.numeric(bp$p.value)
)

kable(bp_df[, 2:3], 
      digits = 4, 
      booktabs = TRUE) |> 
  kable_styling(latex_options = "hold_position", 
                font_size = 9)
```

With a p-value of 0.7439, we fail to reject the null hypothesis of constant error variance, meaning there is no evidence of heteroskedasticity in the model.

### 2.4 Conclusion
Overall, the diagnostic tests support the validity of the OLS assumptions for our first model.
The VIF results (around 1.2–1.3) indicate no multicollinearity, so the regressors are not strongly correlated.
The Shapiro–Wilk test (p = 0.7615) suggests the residuals are approximately normally distributed.
The Breusch–Pagan test (p = 0.7439) shows no evidence of heteroskedasticity, meaning the error variance can be treated as constant.
Taken together, these checks imply that our coefficient estimates and standard inference from this model are generally reliable.

## 3) Modelling tax evasion by billionaires
### 3.1 Creation of a flight-risk score

The aim of this part is to establish, for each billionaire the risk of flight from their current national territory if a tax is introduced. We will therefore create in this section a flight risk score, based on four distinct component :

A tax burden score between 0 and 1 calculated according to the following standardization : $\frac{(\text{tax in country i})- (\text{min tax})}{(\text{max tax})- (\text{min tax})}$. This score is thus normalized between 0 (minimum taxation) and 1 (maximum taxation). We consider that higher taxation would encourage more tax evasion.  

Whether the billionaire's main activity belongs to an industry considered to be internationally mobile. Among the industries in the original dataset, we have identified the following as mobile : "Diversified financial",  "Technology-Computer", "Hedge funds", "Private equity/leverage buyout",      "Venture capital", "Media", "Money Management". Belonging to one of this industry gives a score of 1 and 0 otherwise as we believe that working in such a sector would facilitate mobility for the billionaire. 

The source of the billionaire's wealth, with a score of 1 if "self-made" and 0 if "inherited". This score is used to capture the idea that inherited wealth is more likely to be accompanied by a strong attachment to the territory of origin. The idea is that inherited wealth tends to involve real estate and historically established industries, which provide incentives to stay put. 

Proximity to a tax haven. Based on the official definition, we first drew up the following list of tax havens: "Switzerland", "Singapore", "Hong Kong", "Liechtenstein", "Monaco", "Luxembourg", "United Arab Emirates" and "Bermuda". We then identified the countries considered close enough to a tax haven to provide billionaires with a significant incentive to relocate. To do this, we took into account membership of the Schengen area for European countries (where mobility to the border of the tax haven is facilitated, whether for Switzerland, Luxembourg or Liechtenstein) and the sharing of a coastline with the same ocean for Asian and Caribbean tax havens. This gives us a score of 1 if close, 0 otherwise.

```{r}
#| label: Creation of the 4 scores
#| echo: false
#| message: false
#| warning: false


# Sectors considered "mobile"

mobile_sectors <- c("Diversified financial", "Technology-Computer", "Hedge funds", "Private equity/leverage buyout", "Venture capital", "Media", "Money Management")

# List of tax havens/refuges

tax_havens <- c(
  "Switzerland", "Singapore", "Hong Kong", "Liechtenstein",
  "Monaco", "Luxembourg", "United Arab Emirates", "Bermuda"
)

# Calcul of the 4 indicators

flight_risk_index <-
  Billionaires_2014 |>
  left_join(Tax_revenue_reduced |> select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  mutate(
    # tax burden score calculation
    tax_normalized =
      ifelse(tax != 0, (tax - min(tax, na.rm = TRUE)) / (max(tax, na.rm = TRUE) - min(tax, na.rm = TRUE)), 0),
    tax_normalized =
      ifelse(is.na(tax) == TRUE, 0, tax_normalized),
    # mobility score calculation
    mobile_sector = ifelse(industry %in% mobile_sectors, 1, 0),
    # source of wealth index computation
    selfmade = ifelse(is.na(selfmade) == TRUE, "self-made", selfmade),
    is_selfmade = ifelse(selfmade == "self-made", 1, 0),
    is_selfmade = replace_na(is_selfmade, 0),

    # proximity to tax haven index computation
    near_haven = case_when(
      COUNTRY %in% c(
        "France", "Germany", "Italy", "Spain", "United Kingdom",
        "Belgium", "Netherlands", "Austria", "Sweden", "Norway",
        "Portugal", "Denmark", "Czech Republic", "Finland", "Greece",
        "Poland", "Romania", "Cyprus", "Lithuania"
      ) ~ 1,
      COUNTRY %in% c(
        "China", "Japan", "South Korea", "India", "Malaysia",
        "Thailand", "Indonesia", "Taiwan", "Philippines", "Macau", "Vietnam"
      ) ~ 1,
      COUNTRY %in% c(
        "Saudi Arabia", "Kuwait", "Israel", "Turkey", "Egypt",
        "Lebanon", "Oman"
      ) ~ 1,
      COUNTRY %in% c(
        "United States", "Canada", "Brazil", "Mexico", "Colombia",
        "Venezuela", "St. Kitts and Nevis"
      ) ~ 1,
      TRUE ~ 0
    ),
    in_tax_haven = ifelse(COUNTRY %in% tax_havens, 1, 0)
  )
```

For each billionaire, four indicators are calculated, which are then combined to obtain the final flight risk score. For the calculation, we assigned the following coefficients :

$$
\text{Score} = 0.45 \cdot \text{Taxation} + 0.3 \cdot \text{Mobility} + 0.15 \cdot \text{Source of wealth} + 0.1 \cdot \text{Proximity}
$$

Taxation has the highest coefficient because it is theoretically the most significant incentive to leave. Mobility comes next, as it represents the most difficult physical constraint to overcome. The origin of wealth has a less significant coefficient because its impact remains limited due to the complexity of accurately establishing a billionaire's personal ties to their territory. The smallest coefficient is associated with proximity to a tax haven. Despite the physical constraint that distance represents, the economy remains globalized and billionaires have ample means to circumvent the issue.

Tax havens are assigned a default score of 0, and for the sake of clarity, categories are assigned according to the score: "Low" if below 0.3, “Medium” between 0.3 and 0.6, and "High" if above 0.6.

```{r}
#| label: Aggregation of the 4 scores
#| echo: false
#| message: false
#| warning: false

flight_risk_index <-
  flight_risk_index |>
  mutate(
    wealth_bil = networthusbillion / 10, # Conversion of tenths of billions into billions
    flight_risk_score = (tax_normalized * 0.45 +
      mobile_sector * 0.30 +
      is_selfmade * 0.15 +
      near_haven * 0.10),
    flight_risk_score = ifelse(in_tax_haven == 1, 0, flight_risk_score),
    risk_category = case_when(
      flight_risk_score < 0.3 ~ "Low",
      flight_risk_score >= 0.3 & flight_risk_score < 0.6 ~ "Medium",
      flight_risk_score >= 0.6 ~ "High"
    )
  )
```

### 3.2 Conversion of the flight-score into a probability using a logistic transformation

We convert the flight-risk score into an expected flight probability using a logistic transformation. The score is a 0–1 index that ranks individuals by exposure to mobility incentives, but it cannot be read directly as a probability (or as the share of wealth that would relocate). The logistic mapping keeps predicted values bounded between 0 and 1 and introduces a realistic non-linearity: changes in risk have limited effects for low-risk profiles, while the response becomes stronger for high-risk profiles without mechanically implying extreme flight rates.

Because we do not observe actual relocation decisions in our data, we calibrate (rather than estimate) the logistic parameters $(\alpha)$ and $(\beta)$ transparently using two anchor points from the score distribution. 
We define $(p_i=\text{plogis}(\alpha+\beta,\text{score}*i))$, and impose that the 10th percentile of the score $((s*{10}))$ corresponds to a low expected flight rate $(p_{10}=2\%)$, and the 90th percentile $((s_{90}))$ correspond to a higher but still plausible rate $(p_{90}=25\%)$. This "moderate mobility" calibration pins down $(\alpha)$ and $(\beta)$ uniquely and yields an average implied flight probability of about 10.8\% in the sample. 

We could test the model according to the type of calibration we want :

- "Moderate" calibration: (2% → 25%) used here. We calibrate the logistic transformation by requiring that individuals in the 10th percentile of the score have an expected leakage of approximately 2%, and those in the 90th percentile have an expected leakage of 25%. This gives an average probability of approximately 10.8%
- "Low" calibration: (1% → 15%)
- "High" calibration: (3% → 35%)


We used AI assistance to refine and validate the implementation of the logit-based calibration code. 


```{r}
#| label: Conversion using the logistic function
#| echo: false
#| message: false
#| warning: false


logit <- function(p) log(p / (1 - p)) # Define the logit function: logit(p) = log(p / (1 - p))
# two anchor points from the empirical distribution of the score:
s_low <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.10, na.rm = TRUE)) #    s_low  = 10th percentile of flight_risk_score (low-risk profiles)
s_high <- as.numeric(quantile(flight_risk_index$flight_risk_score, 0.90, na.rm = TRUE)) #    s_high = 90th percentile of flight_risk_score (high-risk profiles)

# Target flight probabilities at these anchors (moderate mobility scenario):
# p_low  is the expected flight probability at s_low
# p_high is the expected flight probability at s_high
p_low <- 0.02 # ~2% expected flight at the low-risk (10th percentile) score
p_high <- 0.25 # ~25% expected flight at the high-risk (90th percentile) score

# Solve for beta and alpha in:
#    p(score) = plogis(alpha + beta * score)
# Using the logit transform:
#    logit(p_low)  = alpha + beta * s_low
#    logit(p_high) = alpha + beta * s_high
# Subtracting the two equations gives beta, then alpha follows.
beta <- (logit(p_high) - logit(p_low)) / (s_high - s_low)
alpha <- logit(p_low) - beta * s_low

# Compute the implied flight probability for each billionaire:
flight_risk_index <- 
  flight_risk_index |>
  mutate(
    flight_prob = ifelse(flight_risk_score!=0, plogis(alpha + beta * flight_risk_score), 0)
  )
```

### 3.3 Analysis of the results

Now that we have converted the flight-risk score into a calibrated flight probability (flight_prob), we aggregate the results at the country level. This allows us to compare countries in terms of (i) billionaire concentration, (ii) total billionaire wealth, and (iii) both the average risk index and the average implied mobility (probability), focusing on countries with a sufficiently large billionaire sample.

```{r}
#| label: tbl-country-risk
#| tbl-cap: "Top Country Risk Analysis"
#| echo: false
#| message: false
#| warning: false

country_risk <-
  flight_risk_index |>
  summarise(
    nb_billionaires = n(), # number of billionaires
    total_wealth = sum(wealth_bil, na.rm = TRUE), # total wealth ($bn), unit-corrected
    avg_flight_risk = mean(flight_risk_score, na.rm = TRUE), # average risk score (index)
    avg_flight_prob = mean(flight_prob, na.rm = TRUE), # average risk score (prob)
    nb_high_risk = sum(risk_category == "High", na.rm = TRUE), # average calibrated flight probability
    pct_high_risk = (nb_high_risk / n()) * 100, # % of high-risk individuals
    tax = mean(tax, na.rm = TRUE), # average tax proxy (same within a country)
    .by = COUNTRY
  ) |>
  arrange(desc(avg_flight_risk))

top_country_risk <-
  country_risk |>
  filter(nb_billionaires > 10) |>
  slice_max(order_by = avg_flight_prob, n = 10)

top_country_risk |> 
  select(1, 2, 5) |> 
  kable(format = "latex",
        booktabs = TRUE,
        longtable = TRUE,
        digits = 2)

top_country_risk

```

The table aggregates the flight-risk indicators at the country level for countries with more than ten billionaires. It reports the average flight-risk score (our composite index) and the average implied flight probability (from the calibrated logistic mapping). 

The United Kingdom, Israel, and Sweden appear at the top with relatively high average probabilities, which fits an interpretation where a larger share of fortunes is connected to internationally mobile activities (finance, tech, media) and where cross-border relocation is comparatively frictionless. Malaysia and Thailand also rank relatively high despite smaller billionaire populations, suggesting a composition effect: even with fewer individuals, the billionaire profiles present there may be more concentrated in footloose sectors, pushing up the average mobility indicator.

France sits in an intermediate position: its average flight probability is meaningful but below the top countries, consistent with the idea that part of French billionaire wealth is tied to established domestic firms and less mobile assets that anchor residence and business operations. The United States is a useful contrast. Despite having by far the largest number of billionaires, its average flight probability is only moderate. This supports a “home-market” mechanism: deep domestic opportunities and large market scale can reduce the incentive to relocate on average, even though there may still be a sizable high-mobility minority in absolute numbers.


```{r}
#| label: Basic analysis plots
#| echo: false
#| message: false
#| warning: false

plot_distrib <- ggplot(flight_risk_index, aes(x = flight_prob))+
  geom_density()+
  labs(
    title = "Distribution of the flight probability",
    x = "Probability"
  ) +
  theme_clean(base_size = 10)

plot_region <- ggplot(flight_risk_index, aes(x = flight_prob))+
  geom_histogram()+
  facet_wrap(~selfmade)+
  labs(
    title = "Distribution of the flight probability by source of wealth",
    x = "Probability"
  ) +
  theme_clean(base_size = 10)


plot_region / plot_distrib

```
In the top panel, self-made billionaires display higher predicted mobility than inherited ones. Inherited wealth is concentrated at very low flight probabilities, consistent with fortunes that are more anchored in domestic assets and long-established family structures, which make relocation costlier. By contrast, the self-made distribution has a thicker right tail, indicating more individuals with medium-to-high flight probabilities. This fits an economic interpretation where self-made wealth is more often tied to internationally scalable businesses and financial assets that are easier to reorganize across borders.

The bottom density plot summarizes the overall distribution of flight probabilities in the sample. It is strongly right-skewed: most individuals cluster at low probabilities, while a smaller group forms a long tail toward higher values. This matters because it implies heterogeneity in behavioral responses? Indeeed, the flight risk is not evenly spread across billionaires. The density also appears multi-peaked, which is consistent with the index being built from discrete components (mobile sector, proximity, self-made status) combined with a continuous tax term: different “profiles” of billionaires naturally bunch at different probability ranges.

Overall, these figures suggest that mobility risk is not uniform across billionaires. Most individuals cluster at low predicted flight probabilities, especially among inherited fortunes, but there is a clear right tail driven largely by self-made profiles. In other words, the potential response to a wealth tax is concentrated in a smaller, more mobile subset rather than spread evenly across the entire billionaire population.

## 4) Scenario testing : the impact of new taxation 

In this section, we move from descriptive evidence to policy evaluation. Building on our flight-risk framework, we simulate the expected fiscal impact of introducing a new 2% billionaire wealth tax under alternative institutional settings. The goal is to quantify how different degrees of international coordination affect wealth mobility, the erosion of the tax base, and ultimately the amount of revenue that can be effectively collected.

We define a set of nested country coalitions to represent progressively broader degrees of international coordination. We start with the G7 and the G20, which capture coordination among major advanced economies and among the largest global economies, respectively. To extend the analysis beyond these institutional groups, we also build a broader “G20 + North” coalition by adding all countries classified as part of the Global North in our dataset, reflecting a wider agreement among high-income jurisdictions that host a large share of the taxable base. Finally, we consider a “global” benchmark that includes all countries in our sample, which serves as an upper bound on coordination. This tiered structure allows us to assess how expanding participation changes expected mobility incentives and the effective fiscal yield of the tax.

```{r}
#| label: Define G7 and G20 members
#| message: false
#| warning: false
#| echo: false

G7 <- c(
  "United States", "United Kingdom", "Germany", "France",
  "Italy", "Canada", "Japan"
)

G20 <- c(
  G7, "China", "India", "Brazil", "Russia", "Australia",
  "South Korea", "Mexico", "Indonesia", "Turkey", "Saudi Arabia",
  "Argentina", "South Africa"
)

NORTH <- flight_risk_index |>
  filter(north == 1) |>
  distinct(COUNTRY) |>
  pull(COUNTRY)

G20_North <- unique(c(G20, NORTH)) #Coalition 4 = G20 + North countries 

GLOBAL <- unique(flight_risk_index$COUNTRY) #All countries in our sample
```

We then implement a revenue simulation that converts our calibrated individual flight probability into fiscal outcomes under each policy setting. For each billionaire, the baseline flight probability (`flight_prob`) is scaled by a scenario-specific multiplier—higher under unilateral action and lower under broader coordination—and capped at 1 to avoid probabilities above 100%. With a 2% tax rate on billionaire wealth, the simulation computes the potential revenue assuming no flight, the amount of wealth expected to relocate, the remaining taxable base, and the resulting actual revenue. The difference between potential and actual revenue gives the revenue loss attributable to flight, and their ratio provides a collection-efficiency measure. More explicitely, it is defined as the share of the theoretical tax yield that is effectively collected once flight is taken into account. It serves as a summary indicator of tax-base erosion: higher values imply limited leakage and stronger enforceability of the policy, whereas lower values indicate substantial revenue losses due to mobility.

```{r}
#| message: false
#| warning: false
#| echo: false
## CHUNK 2.2 : Fonction de simulation des recettes fiscales
# -----------------------------------------------------------------------------
# Cette fonction calcule les recettes d'une taxe en tenant compte de la fuite
# Elle prend en entrée un coefficient d'ajustement de la probabilité de fuite

simulate_tax_scenario <- function(data, scenario_name, flight_multiplier) {
  # flight_multiplier : ajuste la probabilité de fuite
  # = 1.0 pour fuite normale, < 1.0 pour fuite réduite (coordination)

  results <- data |>
    mutate(
      effective_flight_prob = flight_prob * flight_multiplier, # Probabilité effective de fuite dans ce scénario
      effective_flight_prob = pmin(effective_flight_prob, 1), # Max = 1
      initial_wealth = wealth_bil,
      fled_wealth = initial_wealth * effective_flight_prob, # Richesse qui fuit à l'étranger
      taxable_wealth = initial_wealth - fled_wealth, # Richesse restante (qui sera taxée)
      potential_revenue = initial_wealth * 0.02, # Recettes potentielles (si aucune fuite)
      actual_revenue = taxable_wealth * 0.02, # Recettes réelles (après fuite)
      lost_revenue = potential_revenue - actual_revenue, # Perte de recettes due à la fuite
    )


  # Résumé global
  global_summary <- results |>
    summarise(
      scenario = scenario_name,
      total_billionaires = n(),
      avg_flight_prob = mean(effective_flight_prob) * 100,
      total_wealth_billion = sum(initial_wealth),
      fled_wealth_billion = sum(fled_wealth),
      potential_revenue_billion = sum(potential_revenue),
      actual_revenue_billion = sum(actual_revenue),
      lost_revenue_billion = sum(lost_revenue),
      collection_efficiency = (actual_revenue_billion / potential_revenue_billion) * 100
    )

  # Résumé par pays (top contributeurs)
  country_summary <- results |>
    group_by(COUNTRY) |>
    summarise(
      scenario = scenario_name,
      nb_bill = n(),
      potential = sum(potential_revenue),
      actual = sum(actual_revenue),
      lost = sum(lost_revenue)
    ) |>
    arrange(desc(actual))

  return(list(global = global_summary, by_country = country_summary))
}
```



```{r}

#| label: Test de scénarios
#| echo: false
#| message: false
#| warning: false

sc_fr <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY == "France"), "France seule", 1.5)
sc_g7 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G7), "Coordination G7", 0.9)
sc_g20 <- simulate_tax_scenario(flight_risk_index |> filter(COUNTRY %in% G20), "Coordination G20", 0.7)
sc_g20_north <- simulate_tax_scenario(
  flight_risk_index |> filter(COUNTRY %in% G20_North),
  "Coordination élargie (G20 + North)",
  0.55
)

sc_global <- simulate_tax_scenario(
  flight_risk_index |> filter(COUNTRY %in% GLOBAL),
  "Coordination quasi-globale",
  0.40
)


summary_table_all <- bind_rows(sc_fr$global, sc_g7$global, sc_g20$global, sc_g20_north$global, sc_global$global) |>
  transmute(
    Scenario = scenario,
    `# billionaires` = total_billionaires,
    `Expected flight rate (%)` = round(avg_flight_prob, 1),
    `Potential revenue ($bn)` = round(potential_revenue_billion, 3),
    `Actual revenue ($bn)` = round(actual_revenue_billion, 3),
    `Lost revenue ($bn)` = round(lost_revenue_billion, 3),
    `Collection efficiency (%)` = round(collection_efficiency, 1)
  )

summary_table_all
```
The scenario table highlights a clear policy trade-off between the size of the tax base and the degree of international coordination. Moving from a unilateral French tax to broader coalitions (G7, G20, G20+North, quasi-global) mechanically increases the number of covered billionaires and the potential revenue, because more individuals and more wealth fall under the tax. At the same time, stronger coordination reduces the expected flight response: the implied average flight rate falls monotonically from 18.7% (France only) to 4.0% (quasi-global). This captures the key mechanism of coordination: by reducing the set of attractive “outside options,” coordination lowers the elasticity of the tax base to the policy.

In levels, potential revenue rises from $3.66bn (France) to $104.25bn (quasi-global). However, revenue losses due to flight do not increase monotonically with the coalition size. Losses peak around the intermediate coalitions (G7/G20) and then decline as coordination becomes broader (G20+North and quasi-global). Economically, this is intuitive: when the coalition expands a lot and coordination is stronger, the reduction in flight probabilities dominates the larger tax base, so less wealth is expected to escape. This is also reflected in collection efficiency, which improves steadily from 78.4% under unilateral action to 95.9% under quasi-global coordination. The key message is that coordination increases the effective yield of the wealth tax by compressing tax-base erosion, especially once participation becomes sufficiently broad.

```{r}
#| message: false
#| warning: false
#| echo: false

loss_plot_df <- summary_table_all %>%
  mutate(
    Scenario = factor(
      Scenario,
      levels = c("France seule", "Coordination G7", "Coordination G20",
                 "Coordination élargie (G20 + North)", "Coordination quasi-globale")
    ),
    # useful derived metric to interpret shape independent of scale
    `Loss rate (%)` = 100 * (`Lost revenue ($bn)` / `Potential revenue ($bn)`)
  )

ggplot(loss_plot_df, aes(x = Scenario, y = `Loss rate (%)`, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Tax-base erosion declines with stronger coordination",
    x = NULL,
    y = "Lost revenue as % of potential revenue"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```




```{r}
#| label: Top 5 contributeurs à la perte
#| message: false
#| warning: false
#| echo: false
top_loss_fr <- flight_risk_index |>
  filter(COUNTRY == "France") |>
  mutate(
    p_france = pmin(flight_prob * 1.5, 1),
    potential_rev = wealth_bil * 0.02,
    actual_rev = wealth_bil * (1 - p_france) * 0.02,
    lost_rev = potential_rev - actual_rev
  ) |>
  arrange(desc(lost_rev)) |>
  select(name, wealth_bil, p_france, lost_rev, industry, selfmade) |>
  head(10)

top_loss_fr
```

```{r}
#| message: false
#| warning: false
#| echo: false

fr_detail <- flight_risk_index |>
  filter(COUNTRY == "France") |>
  mutate(
    eff = pmin(flight_prob * 1.5, 1),
    lost = wealth_bil * 0.02 * eff
  ) |>
  arrange(desc(wealth_bil)) |>
  mutate(
    cum_lost_share = cumsum(lost) / sum(lost),
    rank = row_number(),
    pct = rank / n()
  )

# Part des pertes expliquée par le top 10% (les plus riches)
fr_detail |>
  summarise(share_lost_top10 = max(cum_lost_share[pct <= 0.10]))
```
