---
title: "Final_phase"
format: pdf
---


```{r}
#| label: Data cleaning
#| message: false
#| warning: false
#| echo: false

# Library import

library(readxl) # To open the billionaires dataset
library(tidyverse) # To clean, modify and analyze our data
library(ggplot2) # For graphical representation
library(stringr) # To clean the data
library(dplyr)
library(maps)
library(viridis)
library(ggrepel)
library(broom)
library(scales)
library(patchwork)
library(ggthemes)
library(car) #vif
library(lmtest) #heteroscedasticity 

# Data import

Billionaires <- read_delim("Database_project/Billionaires_dataset.csv",
                           delim = ";", escape_double = FALSE, trim_ws = TRUE)

IMF <- read_csv("Database_project/IMF_GDP_PPP.csv")

FDI <- read_csv("Database_project/FDI_WB.csv", skip = 3)

Tax_revenue <- read_csv("Database_project/Tax_revenue.csv", skip = 3)

Capital_formation <- read_csv("Database_project/Capital_formation.csv", skip = 3
)

Population <- read_csv("Database_project/Population.csv", skip = 3)


# Cleaning billionaires data set

useless <- ",;.-()"

Billionaires <-
  Billionaires |>
  select(c(1:22)) |> # removing useless variables
  mutate(name = str_remove_all(name, useless)) |> # standardization of names
  rename(COUNTRY = citizenship)

Billionaires_1996 <-
  Billionaires |>
  filter(year == "1996")

Billionaires_2001 <-
  Billionaires |>
  filter(year == "2001")

Billionaires_2014 <-
  Billionaires |>
  filter(year == "2014")

Billionaires_wide <-
  Billionaires |>
  mutate(year = as.character(year)) |>
  pivot_wider(
    id_cols = c(
      name, COUNTRY, countrycode,
      selfmade, typeofwealth, gender,
      industry, IndustryAggregates, region, north,
      politicalconnection, founder, generationofinheritance,
      sector, company, relationshiptocompany,
      foundingdate, companytype
    ),
    names_from = year,
    values_from = c(rank, networthusbillion, age),
    names_sep = "_"
  )


# Cleaning the IMF dataset

Country <-
  Billionaires |>
  distinct(COUNTRY)

liste_pays <- c(
  "Algeria", "Angola", "Argentina", "Australia", "Austria", "Bahrain",
  "Belgium", "Bermuda", "Brazil", "Canada", "Chile", "China", "Colombia",
  "Cyprus", "Czech Republic", "Denmark", "Ecuador", "Egypt", "Finland",
  "France", "Georgia", "Germany", "Greece", "Guernsey", "Hong Kong", "India",
  "Indonesia", "Ireland", "Israel", "Italy", "Japan", "Kazakhstan", "Kuwait",
  "Lebanon", "Liechtenstein", "Lithuania", "Macau", "Malaysia", "Mexico",
  "Monaco", "Morocco", "Nepal", "Netherlands", "New Zealand", "Nigeria",
  "Norway", "Oman", "Peru", "Philippines", "Poland", "Portugal", "Romania",
  "Russia", "Saudi Arabia", "Singapore", "South Africa", "South Korea",
  "Spain", "St. Kitts and Nevis", "Swaziland", "Sweden", "Switzerland",
  "Taiwan", "Tanzania", "Thailand", "Turkey", "Uganda", "Ukraine",
  "United Arab Emirates", "United Kingdom", "United States",
  "Venezuela", "Vietnam"
)

map_IMF <- c(
  "Bahrain, Kingdom of" = "Bahrain",
  "China, People's Republic of" = "China",
  "Egypt, Arab Republic of" = "Egypt",
  "Hong Kong Special Administrative Region, People's Republic of China" = "Hong Kong",
  "Kazakhstan, Republic of" = "Kazakhstan",
  "Liechtenstein, Principality of" = "Liechtenstein",
  "Lithuania, Republic of" = "Lithuania",
  "Macao Special Administrative Region, People's Republic of China" = "Macau",
  "Netherlands, The" = "Netherlands",
  "Poland, Republic of" = "Poland",
  "Russian Federation" = "Russia",
  "Korea, Republic of" = "South Korea",
  "Eswatini, Kingdom of" = "Swaziland",
  "Taiwan Province of China" = "Taiwan",
  "Tanzania, United Republic of" = "Tanzania",
  "Türkiye, Republic of" = "Turkey",
  "Venezuela, República Bolivariana de" = "Venezuela"
)

IMF <-
  IMF |>
  mutate(COUNTRY = recode(COUNTRY, !!!map_IMF))

IMF <-
  IMF |>
  filter(COUNTRY %in% liste_pays)

IMF_reduced <-
  IMF |>
  select(COUNTRY, `1996`, `2000`, `2014`)

# Cleaning the WB datasets

WB_names <-
  FDI |>
  distinct(`Country Name`)

map_WB <- c(
  "Czechia" = "Czech Republic",
  "Egypt, Arab Rep." = "Egypt",
  "Hong Kong SAR, China" = "Hong Kong",
  "Macao SAR, China" = "Macau",
  "Russian Federation" = "Russia",
  "Korea, Rep." = "South Korea",
  "Venezuela, RB" = "Venezuela",
  "Viet Nam" = "Vietnam"
)

FDI <-
  FDI |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

FDI <-
  FDI |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

FDI_reduced <-
  FDI |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Tax_revenue <-
  Tax_revenue |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Tax_revenue <-
  Tax_revenue |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Tax_revenue_reduced <-
  Tax_revenue |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Capital_formation <-
  Capital_formation |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Capital_formation <-
  Capital_formation |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Capital_formation_reduced <-
  Capital_formation |>
  select(COUNTRY, `1996`, `2000`, `2014`)

Population <-
  Population |>
  mutate(`Country Name` = recode(`Country Name`, !!!map_WB))

Population <-
  Population |>
  filter(`Country Name` %in% liste_pays) |>
  rename(COUNTRY = `Country Name`)

Population_reduced <-
  Population |>
  select(COUNTRY, `1996`, `2000`, `2014`)

```

# Introduction
## How does taxing billionaire wealth affect economic performance, inequality, and national competitiveness in an open economy?

Billionaire taxation has emerged as one of the most polarizing economic questions of our decade. In October 2024, French economist Gabriel Zucman presented to the G20 a proposal for a minimum 2% tax on fortunes exceeding one billion dollars, potentially generating $200-250 billion annually worldwide. Yet the debate divides deeply: while supporters see it as essential for fiscal equity and financing public services in an era of unprecedented wealth concentration (the top 1% now holds nearly 45% of global wealth), opponents fear massive capital flight toward more lenient jurisdictions. France perfectly illustrates this tension—with public debt exceeding 110% of GDP and approximately 40 billionaires whose combined wealth exceeds €500 billion. A 2% tax could theoretically yield €10 billion annually, but the country's experience with its abolished wealth tax (ISF) fuels concerns about fiscal exodus. The Zucman tax thus presents not merely a fiscal tool but a test of our ability to balance fiscal justice imperatives with economic competitiveness in regulating 21st century capitalism.


## Data sets description

- The Peterson Institute for International Economics (2016): A dataset of global billionaires featuring 23 variables-such as citizenship, net worth, and origin of wealth-used to track the demographic and financial characteristics of the world's wealthiest individuals.
- National Fiscal Data (Tax Proxy): A dataset focusing on government tax revenue as a percentage of total revenue, utilized as a proxy for a country's fiscal pressure and level of taxation in response to the limited availability of direct tax rates in billionaire-heavy jurisdictions.
- The World Bank (Macroeconomic Indicators): A collection of global economic metrics-including Foreign Direct Investment (FDI), Gross Capital Formation (% of GDP), Total Population, and GDP in PPP—used to control for market scale, investment climate, and national productivity.

# PARTIE I : LES DÉTERMINANTS DE LA RICHESSE ET DE LA LOCALISATION DES MILLIARDAIRES

## 1) State of the world of billionaires in 2014

### 1.1 Global mapping of billionaires

```{r}
#| message: false
#| warning: false
#| echo: false

# Aggregation by country: number and total wealth

billionaires_by_country <- Billionaires_2014 |>
  group_by(COUNTRY) |> # Group data by country
  summarise(
    nb_billionaires = n(), # Count the number of billionaires per country
    total_wealth = sum(networthusbillion, na.rm = TRUE), # Sum total wealth, removing NA
    avg_wealth = mean(networthusbillion, na.rm = TRUE) # Calculate average wealth per billionaire
  ) |>
  ungroup() # Remove grouping structure

# Preparing cartographic data

world_map <- map_data("world") # Load world map coordinates from maps package

# Mapping country names to match map_data format

country_mapping <- c(
  "United States" = "USA",
  "United Kingdom" = "UK",
  "Russia" = "Russia",
  "South Korea" = "South Korea",
  "Czech Republic" = "Czech Republic"
)

billionaires_map <- billionaires_by_country |>
  mutate(region = ifelse(COUNTRY %in% names(country_mapping), # Check if country needs mapping
    country_mapping[COUNTRY], # Apply mapping if found
    COUNTRY
  )) # Keep original name otherwise

# Join billionaire data with map data

world_billionaires <- world_map |>
  left_join(billionaires_map, by = "region") # Merge map and billionaire data by region name

# Map: Number of billionaires by country

map_count <- ggplot() +
  geom_polygon(
    data = world_map, # Draw all countries as base layer
    aes(x = long, y = lat, group = group), # Define coordinates and grouping
    fill = "grey90", color = "white", linewidth = 0.1
  ) + # Draw only countries with billionaires
  geom_polygon(
    data = world_billionaires |> filter(!is.na(nb_billionaires)), # Color by number of billionaires
    aes(x = long, y = lat, group = group, fill = nb_billionaires), # White borders between countries
    color = "white", linewidth = 0.1
  ) +
  scale_fill_viridis(
    option = "plasma",
    name = "Number of\nbillionaires",
    trans = "log10", # Apply logarithmic transformation
    breaks = c(1, 5, 10, 50, 100, 500),
    direction = -1
  ) +
  theme_void() + # Remove axes and background for clean map
  labs(title = "Global distribution of billionaires in 2014") +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(map_count)
```
For this specific map, we used AI assistance to help refine the R code and generate a clearer, more visual representation. We already had the initial intuition and the base approach, but we chose this mapping option because it is easier to read and more informative than a standard “classic” ggplot output.

This map shows a strong concentration of billionaires in a small number of large, high-income economies. The United States clearly stands out as the main hub, while Western Europe and parts of East Asia form a second cluster. Russia and India also appear as major centers, highlighting the role of large domestic markets and resource wealth. In contrast, most of Sub-Saharan Africa and many smaller or lower-income countries show few or none, underlining how billionaire wealth tends to align with economic scale, financial development, and the presence of major business hubs.


### 1.2 Caracteristics and origin of billionaires' wealth 

```{r}
#| message: false
#| warning: false
#| echo: false

Sectors <-
  Billionaires_2014 |>
  group_by(sector) |>
  summarise(
    nombre_milliardaires = n(),
    nb_self_made = sum(selfmade == "self-made"),
    nb_inherited = sum(selfmade == "inherited")
  ) |>
  mutate(
    pct_self_made = (nb_self_made / nombre_milliardaires) * 100,
    pct_inherited = (nb_inherited / nombre_milliardaires) * 100
  ) |>
  slice_max(n = 10, order_by = nombre_milliardaires) |>
  pivot_longer(
    cols = c(nb_self_made, nb_inherited),
    names_to = "origine",
    values_to = "count"
  ) |>
  mutate(origine = ifelse(origine == "nb_self_made", "Self-made", "Inherited"))


ggplot(Sectors, aes(x = reorder(sector, nombre_milliardaires), y = count, fill = origine)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Distribution of billionaires by sector and origin of wealth in 2014",
    x = "Sector",
    y = "Total number of billionaires",
    fill = "Type of welath"
  ) +
  theme_clean(base_size = 10)
```

This first graph illustrates the distribution of billionaires across various industries, highlighting that real estate is the most prominent sector for amassing such wealth, followed by retail and construction. A significant trend across almost all sectors is the clear dominance of self-made individuals over those who inherited their fortunes. This disparity is particularly pronounced in modern industries like technology, software, and hedge funds, where inherited wealth is virtually non-existent, whereas more traditional sectors like oil and banking show a relatively higher presence of generational wealth.


```{r}
#| message: false
#| warning: false
#| echo: false


Source <-
  Billionaires |>
  drop_na(selfmade) |>
  summarise(n = n(), .by = c(selfmade, year))

ggplot(Source, aes(y = factor(year), x = n, fill = selfmade)) +
  geom_col() +
  labs(
    title = "Source of wealth by year",
    y = "Year",
    x = "Number",
    fill = "Origin"
  ) +
  theme_clean(base_size = 10)
```

This graph tracks the evolution of billionaire wealth origins from 1996 to 2014, revealing a dramatic surge in the total number of billionaires globally, which increased from approximately 400 to over 1,600. While the billionaire population was much smaller and more evenly divided between inherited and self-made origins in 1996, the landscape shifted significantly by 2014. The expansion of this demographic over the eighteen-year period has been overwhelmingly driven by self-made entrepreneurs, who now constitute the vast majority of the total, indicating a global shift toward new wealth creation rather than the simple maintenance of inherited estates.

## 2) First model
```{r}
#| label: First model
#| message: false
#| warning: false
#| echo: false

# Data preparation

reg_data <- Billionaires_2014 |>
  group_by(COUNTRY) |>
  summarise(nb_bill = n()) |>
  left_join(Tax_revenue_reduced |>
              select(COUNTRY, tax = `2014`), by = "COUNTRY") |>
  left_join(IMF_reduced |>
              select(COUNTRY, gdp = `2014`), by = "COUNTRY") |>
  left_join(FDI_reduced |>
              select(COUNTRY, fdi = `2014`), by = "COUNTRY") |>
  left_join(Capital_formation_reduced |>
              select(COUNTRY, capital = `2014`), by = "COUNTRY") |>
  mutate(
    log_bill = log(nb_bill + 1),
    log_gdp = log(gdp)
  ) |>
  filter(!is.na(tax) & !is.na(gdp) & !is.na(fdi) & !is.na(capital))

# First model

modele <- lm(log_bill ~ tax + log_gdp + fdi + capital, data = reg_data)
summary(modele)


```

### 2.1 Multicolinearity analysis

We here test for multicollinearity among the explanatory variables using the Variance Inflation Factor (VIF).

```{r}
#| label: Multicolinearity analysis
#| message: false
#| warning: false
#| echo: false
vif_df <- data.frame(
  Variable = names(vif(modele)),
  VIF = as.numeric(vif(modele))
)  
vif_df
```

The VIF values (≈1.19–1.34) are close to 1, indicating no meaningful multicollinearity among the regressors, so coefficient significance is unlikely to be driven by correlated explanatory variables.


### 2.2 Normality test for residuals

Here, we test whether our data are distributed according to a normal distribution. To do this, we calculate the Shapiro-Wilk normality test statistics. We obtain the following results :


```{r}
#| label: Normality test for residuals
#| message: false
#| warning: false
#| echo: false

shapiro_test <- shapiro.test(residuals(modele))

shapiro_test

```

### 2.3 Heteroscedasticity 

We here test for heteroscedasticity using the Breusch–Pagan test to check whether the error variance is constant. 
```{r}
#| label: Heteroscedasticy 
#| message: false
#| warning: false
#| echo: false

bp<-bptest(modele)


bp_df <- data.frame(
  Test = "Studentized Breusch–Pagan",
  Test_statistic = as.numeric(bp$statistic),
  p_value = as.numeric(bp$p.value)
)

bp_df
```

With a p-value of 0.7439, we fail to reject the null hypothesis of constant error variance, meaning there is no evidence of heteroskedasticity in the model.


```{r}
# =============================================================================
# TEST 6 : AJOUT DE VARIABLES DE CONTRÔLE (Robustesse)
# =============================================================================

cat("\n=== TEST 6 : ROBUSTESSE AVEC CONTRÔLES SUPPLÉMENTAIRES ===\n\n")

# Ajout de la population
reg_data_pop <- reg_data %>%
  left_join(Population_reduced %>% select(COUNTRY, pop = `2014`), by = "COUNTRY") %>%
  mutate(log_pop = log(pop)) %>%
  filter(!is.na(pop))

model_with_pop <- lm(log_bill ~ tax + log_gdp + fdi + capital + log_pop, 
                     data = reg_data_pop)

cat("Modèle avec population:\n")
print(summary(model_with_pop)$coefficients)

cat("\nComparaison coefficient tax:\n")
cat("Sans population:", round(coef(modele)["tax"], 4), 
    "(p =", round(summary(modele)$coefficients["tax", "Pr(>|t|)"], 4), ")\n")
cat("Avec population:", round(coef(model_with_pop)["tax"], 4), 
    "(p =", round(summary(model_with_pop)$coefficients["tax", "Pr(>|t|)"], 4), ")\n\n")

```

```{r}

# =============================================================================
# TEST 7 : EFFET SEUIL DE LA TAXATION
# =============================================================================

cat("\n=== TEST 7 : EXISTE-T-IL UN EFFET SEUIL ? ===\n\n")

# Test avec différents seuils
thresholds <- c(15, 20, 25)

for(threshold in thresholds) {
  reg_data$high_tax <- ifelse(reg_data$tax > threshold, 1, 0)
  model_threshold <- lm(log_bill ~ high_tax + log_gdp + fdi + capital, data = reg_data)
  
  cat("Seuil =", threshold, "%\n")
  cat("  Coefficient high_tax:", round(coef(model_threshold)["high_tax"], 4), "\n")
  cat("  P-value:", round(summary(model_threshold)$coefficients["high_tax", "Pr(>|t|)"], 4), "\n")
  cat("  Significatif:", ifelse(summary(model_threshold)$coefficients["high_tax", "Pr(>|t|)"] < 0.05, 
                                "OUI ✓", "NON ✗"), "\n\n")
}

```

```{r}
# =============================================================================
# TEST 8 : EFFET NON-LINÉAIRE (Polynôme)
# =============================================================================

cat("\n=== TEST 8 : EFFET NON-LINÉAIRE DE LA TAXATION ===\n\n")

reg_data$tax_sq <- reg_data$tax^2
model_nonlinear <- lm(log_bill ~ tax + tax_sq + log_gdp + fdi + capital, data = reg_data)

cat("Modèle avec tax²:\n")
print(summary(model_nonlinear)$coefficients[1:3, ])

cat("\nTest de significativité conjointe (tax + tax²):\n")
model_without_tax <- lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)
f_test <- anova(model_without_tax, model_nonlinear)
cat("F-statistic:", round(f_test$F[2], 4), "\n")
cat("P-value:", round(f_test$`Pr(>F)`[2], 4), "\n")
cat("Interprétation:", ifelse(f_test$`Pr(>F)`[2] < 0.05, 
                              "La taxation a un effet significatif (linéaire ou non-linéaire) ✓", 
                              "La taxation n'a pas d'effet significatif ✗"), "\n\n")
```



```{r}
# =============================================================================
# VISUALISATIONS DIAGNOSTIQUES
# =============================================================================

cat("\n=== VISUALISATIONS DIAGNOSTIQUES ===\n")

# 1. Résidus vs Fitted
plot(modele, which = 1, main = "Résidus vs Valeurs ajustées")

# 2. Q-Q Plot
plot(modele, which = 2, main = "Q-Q Plot (normalité)")

# 3. Scale-Location
plot(modele, which = 3, main = "Scale-Location (homoscédasticité)")

# 4. Cook's distance
plot(modele, which = 4, main = "Cook's distance (influence)")

# 5. Effet partiel de la taxation
plot(reg_data$tax, residuals(lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)),
     xlab = "Taxation (% PIB)", ylab = "Résidus partiels",
     main = "Effet partiel de la taxation (contrôlant pour autres variables)")
abline(lm(residuals(lm(log_bill ~ log_gdp + fdi + capital, data = reg_data)) ~ 
            reg_data$tax), col = "red")

```


```{r}

# =============================================================================
# SYNTHÈSE FINALE
# =============================================================================

cat("\n\n")
cat("═══════════════════════════════════════════════════════════════════\n")
cat("                    SYNTHÈSE DES TESTS                             \n")
cat("═══════════════════════════════════════════════════════════════════\n\n")

cat("QUESTION : La taxation affecte-t-elle la présence de milliardaires ?\n\n")

cat("RÉSULTATS PRINCIPAUX :\n")
cat("• Coefficient taxation:", round(coef(modele)["tax"], 4), "\n")
cat("• P-value:", round(summary(modele)$coefficients["tax", "Pr(>|t|)"], 4), "\n")

# Correction de la ligne problématique
tax_pval <- summary(modele)$coefficients["tax", "Pr(>|t|)"]
if(tax_pval < 0.05) {
  cat("• Significativité: Significatif à 5%\n")
} else if(tax_pval < 0.1) {
  cat("• Significativité: Significatif à 10% (faiblement)\n")
} else {
  cat("• Significativité: Non significatif\n")
}

cat("• R² ajusté:", round(summary(modele)$adj.r.squared, 4), "\n\n")

cat("TESTS DE ROBUSTESSE :\n")
cat("1. Multicolinéarité (VIF) :", ifelse(max(vif_values) < 5, "✓ OK", "✗ Problème"), "\n")
cat("2. Normalité résidus :", ifelse(shapiro_test$p.value > 0.05, "✓ OK", "✗ Problème"), "\n")
cat("3. Homoscédasticité :", ifelse(bp_pvalue > 0.05, "✓ OK", "✗ Problème"), "\n")
cat("4. Observations influentes :", ifelse(length(influential) == 0, "✓ Aucune", 
                                           paste("✗", length(influential), "détectée(s)")), "\n")
cat("5. Stabilité coefficient :", ifelse(sd(loo_coefs) < 0.01, "✓ Stable", "✗ Instable"), "\n\n")

cat("CONCLUSION :\n")
if(tax_pval < 0.05) {
  cat("La taxation a un effet SIGNIFICATIF mais FAIBLE sur la présence de milliardaires.\n")
  cat("L'effet est", ifelse(coef(modele)["tax"] < 0, "NÉGATIF", "POSITIF"), 
      ": +1% de taxation → ", round((exp(coef(modele)["tax"]) - 1) * 100, 2), 
      "% de variation du nombre de milliardaires.\n")
} else if(tax_pval < 0.1) {
  cat("La taxation a un effet FAIBLEMENT SIGNIFICATIF (p < 0.10).\n")
  cat("L'effet existe mais reste MARGINAL par rapport au PIB et aux autres facteurs.\n")
} else {
  cat("La taxation n'a PAS d'effet statistiquement significatif.\n")
  cat("Les OPPORTUNITÉS ÉCONOMIQUES (PIB, investissements) sont les vrais déterminants.\n")
}

cat("\nIMPLICATIONS POUR LA POLITIQUE PUBLIQUE :\n")
cat("→ La taxation seule ne détermine PAS la localisation des milliardaires\n")
cat("→ La TAILLE du marché (GDP) est le facteur dominant\n")
cat("→ Les craintes de 'fuite fiscale massive' sont peut-être exagérées\n")
cat("→ Une politique fiscale équilibrée est possible sans perdre les ultra-riches\n\n")

cat("═══════════════════════════════════════════════════════════════════\n")
```

